---
title: "RA lessons"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: hide
    css: styles.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# Introduction

These are some notes from working with data for economics research projects, mainly in R and often with spatial data. More or less, things worth considering when constructing datasets, checking results and so on.

```{r load-packages}
pacman::p_load(
  dplyr, labelled, sjlabelled, tidyr, ggplot2, scales, sf,
  here, readr, dtplyr, duckdb, DBI, rnaturalearth
  )

options(scipen = 999)
```

maybe say something about whether the code should be aimed at humans or ai - recommend something that is at least readable and debuggable by humans, even if not the cleanest. maybe use some example of a way that ai would write a code 'efficiently' vs humans - perhaps loop is a good example? or even just the number of messages etc

# Working In R

## Style And Speed

I mainly used `dplyr`, due to a combination of preference for its syntax and the piped structure, and being used to it. 


```{r make-style-example}
example_df <- tibble::tibble(
  iso2 = c("US", "US", "CN", "CN"),
  gdp_t = c(25, 26, 17, 18)
)
```

- highly recommend occasionally checking updates packages that are used often, to know when packages add improved functions. Automatically updating them is somewhat risky, if functions with a given name change functionality
- the new pipe (|>) is nicer and cleaner than the old magrittr `%>%` pipe

see the difference between coding styles:

```{r old-pipe-style, results='hide', class.source='fold-show'}
example_df %>% 
  group_by(iso2) %>%
  summarise(
    avg_gdp = mean(gdp_t)
    ) %>%
  ungroup()
```

```{r native-pipe-style, results='hide', class.source='fold-show'}
example_df |> 
  summarise(
    avg_gdp = mean(gdp_t),
    .by = iso2
    ) 
```
Decide for yourself but I much prefer the latter.

`data.table` is faster, but for most processes the marginal improvement is rather small. 

For heavier processes, one can use `dtplyr`. For smaller processes, it seems better to use `dplyr` regularly, as the setup costs may take longer than the actual calculation.

```{r benchmark-dplyr-dtplyr}
set.seed(23072026)

speed_df <- tibble::tibble(
  iso2 = sample(sprintf("C%03d", 1:200), 3000000, replace = TRUE),
  year = sample(2000:2024, 3000000, replace = TRUE),
  gdp_t = runif(3000000, 1, 100),
  pop_m = runif(3000000, 1, 1500)
)

dplyr_elapsed <- system.time({
  dplyr_result <- speed_df |>
    mutate(
      gdp_pc = gdp_t * 1000000 / pop_m,
      large = gdp_t > 50
    ) |>
    filter(large, year >= 2010) |>
    summarise(
      avg_gdp_pc = mean(gdp_pc),
      total_gdp = sum(gdp_t),
      n = n(),
      .by = c(iso2, year)
      )
})[["elapsed"]]

cat("dplyr elapsed:", round(dplyr_elapsed, 3), "seconds\n")

dtplyr_elapsed <- system.time({
  dtplyr_result <- speed_df |>
    lazy_dt() |>
    mutate(
      gdp_pc = gdp_t * 1000000 / pop_m,
      large = gdp_t > 50
    ) |>
    filter(large, year >= 2010) |>
    summarise(
      avg_gdp_pc = mean(gdp_pc),
      total_gdp = sum(gdp_t),
      n = n(),
      .by = c(iso2, year)
      ) |>
    as_tibble()
})[["elapsed"]]

cat("dtplyr elapsed:", round(dtplyr_elapsed, 3), "seconds\n")
```

`duckdb` is another fast option, especially when the data is too large to comfortably keep in memory. It requires SQL syntax, however.

```{r benchmark-duckdb}
con <- dbConnect(duckdb())
duckdb_register(con, "speed_df", speed_df)

duckdb_elapsed <- system.time({
  duckdb_result <- dbGetQuery(
    con,
    "
    SELECT
      iso2,
      year,
      AVG(gdp_t * 1000000 / pop_m) AS avg_gdp_pc,
      SUM(gdp_t) AS total_gdp,
      COUNT(*) AS n
    FROM speed_df
    WHERE gdp_t > 50
      AND year >= 2010
    GROUP BY iso2, year
    "
  )
})[["elapsed"]]

cat("duckdb elapsed:", round(duckdb_elapsed, 3), "seconds\n")

dbDisconnect(con, shutdown = TRUE)
```

## Functions With The Same Names 

Some R functions from different packages have the identical names. This is the source of some bugs and unexpected behaviour, that is sometimes quite hard to detect. 
The `lag` function is a nice example of that. For ordinary vectors/data frames, `stats::lag` can lead to bugs but might be the default; meanwhile, `dplyr::lag` (or `data.table::shift(type = "lag")`) is clearer and more reliable.

```{r compare-lag-functions}
lag_example <- tibble::tibble(
  year = 2020:2022,
  gdp = c(100, 120, 140)
)

lag_example |>
  mutate(
    `dplyr::lag` = dplyr::lag(gdp),
    `stats::lag` = as.numeric(stats::lag(gdp))
  )
```

Similarly, if the code uses `select` without specifying that it should call `dplyr::select`, while `raster` is called, the code might break given the existence of `raster::select`. Here, at least, there wouldn't be a mysterious bug, unlike the `lag` case). 

Another example is `copy_labels`, which exists in both `sjlabelled` and `labelled`. The two versions do not use the same argument order, and they may also change values while copying labels, so it is worth writing the package name explicitly and checking the documentation.

# Data Construction And Cleaning

When reading data, parse dates explicitly rather than trusting whatever type R guessed. For example, use `as.Date()` or `lubridate` functions, then check that the result looks right; `"01/02/2020"` is not self-explanatory unless we know the date format.

## Joins

Quite often we want to join datasets, using one or more common keys. Before joining, I would usually ask: which rows should stay, is the key unique, do both datasets use the same version of the key, and what should happen to unmatched rows or duplicate column names?

There are several types of joins:

| Join | What it keeps | When to use |
|---|---|---|
| `left_join` | all rows from the first dataset | adding variables to a main dataset |
| `inner_join` | only rows that match in both datasets | keeping only units that exist in both sources |
| `full_join` | all rows from both datasets | checking coverage across sources, less common |
| `right_join` | all rows from the second dataset | same idea as `left_join`, but less convenient |
| `semi_join` | rows from the first dataset that have a match in the second, without adding columns | filtering or quick diagnostics, less common |
| `anti_join` | rows from the first dataset that do not have a match in the second | finding unmatched observations |

I usually like to have a "master" dataset with the units that are used for the final analysis, and then `left_join` other datasets into it. During data construction, `inner_join` and `full_join` can also be useful at times, and I rarely use the others - although `anti_join` is good for debugging.

Another consideration is the relationship between the keys:

| Relationship | Meaning | Example |
|---|---|---|
| one-to-one | each key appears once in both datasets | one row per country in both datasets |
| many-to-one | many rows in the main data match one row in the added data | country-year data joined to one region per country |
| one-to-many | one row in the main data matches many rows in the added data | one country joined to several yearly rows |
| many-to-many | keys repeat on both sides | usually risky unless explicitly intended |

The default in `dplyr` is  `relationship = NULL` which does not enforce a relationship, but gives a warning for many-to-many join that was not stated as such. If we know what the relationship should be, it can be useful to state it explicitly, for example `relationship = "many-to-one"`, so the code fails when the data do not match the assumption.

Another common issue is conflicting columns. If both datasets have `country` and `country_name`, for example, and we join on `country` only, then R may create columns such as `country_name.x` and `country_name.y`. That is sometimes useful for checking differences across sources, but usually it is cleaner to either use explicit suffixes or select away the unnecessary version before joining.

The example below is deliberately small. We have a GDP file and a country-region file, and we want to add GDP to the country-region data.

```{r simple-country-join-example}
country_gdp <- tibble::tibble(
  iso3 = c("USA", "CHN", "IND"),
  country_name = c("United States", "China", "India"),
  gdp_b = c(23000, 18000, 3500)
)

country_regions <- tibble::tibble(
  iso3 = c("USA", "CHN", "IND", "MEX"),
  country_name = c("United States of America", "China", "India", "Mexico"),
  region = c("North America", "East Asia", "South Asia", "Latin America")
)

country_gdp

country_regions
```

Joining by ISO3 keeps Mexico in the master country-region file, but leaves its GDP as missing. This is usually what I want when the first dataset defines the analysis units:

```{r left-join-country-example}
joined_country_regions <- country_regions |>
  left_join(
    country_gdp |> select(-country_name),
    join_by(iso3)
  )

joined_country_regions
```

Using `inner_join()` instead would drop Mexico, because it does not appear in the GDP file:

```{r inner-join-country-example}
country_regions |>
  inner_join(
    country_gdp |> select(-country_name),
    join_by(iso3)
  )
```

Joining by country name looks tempting, but fails here because the United States is written differently across the two datasets:

```{r bad-country-name-join-example}
country_regions |>
  left_join(country_gdp, join_by(country_name))
```

If both datasets contain a column with the same name, R may keep both versions with suffixes. In the HTML, the column names are the main thing to notice here:

```{r join-suffix-example}
country_regions |>
  left_join(
    country_gdp,
    join_by(iso3),
    suffix = c("_regions", "_gdp")
  ) |>
  names()
```

Or, if one version is not needed, select it away before joining. This keeps the resulting data cleaner:

```{r clean-country-join-example}
country_regions |>
  left_join(
    country_gdp |> select(-country_name),
    join_by(iso3)
  )
```

I would usually add a tiny diagnostic table after a join:

```{r country-join-diagnostic-table}
country_regions |>
  summarise(
    rows_before = n(),
    rows_after = nrow(joined_country_regions),
    missing_gdp = sum(is.na(joined_country_regions$gdp_b)),
    unmatched_iso3 = nrow(anti_join(country_regions, country_gdp, join_by(iso3)))
  )
```

The key points are that `left_join` and `inner_join` do not keep the same rows, country names are fragile join keys, and duplicate label columns should be handled deliberately rather than carried through the pipeline by accident.

## Messages And Quick Checks

One thing I have not done enough, but probably should, is adding a few short messages after important construction steps, to make mistakes easier to notice. Adding too many can make the code noisy, but after joins or aggregations they can be useful.

For example, after a join it is often worth printing how many units matched, how many did not, and how many rows have missing values in the new variable. After a `summarise`, one could print how many rows were collapsed or how many groups have missing values. 

```{r join-messages, message=TRUE}
joined_country_regions <- country_regions |>
  left_join(
    country_gdp |> 
      select(-country_name),
    join_by(iso3)
  )

unmatched_country_regions <- country_regions |>
  filter_out(iso3 %in% country_gdp$iso3)

message(
  n_distinct(country_regions$iso3) - nrow(unmatched_country_regions), " of ",
  n_distinct(country_regions$iso3), " country-region units matched to GDP."
)

message(
  nrow(unmatched_country_regions), " country-region units did not match to GDP."
)

message(
  sum(is.na(joined_country_regions$gdp_b)), " rows have missing GDP after the join."
)
```

## Aggregates

In some datasets there may be aggregates (e.g., "Middle East" in a country-level dataset). In some datasets there are variables that refer to those, but not always, and the names of the 'aggregate' countries might be different across different datasets (e.g., in two different datasets of FAOSTAT, there is in one dataset (https://www.fao.org/faostat/en/#data/FBSH) an aggregated region of 'Low Income Food Deficit Countries', but in another (https://www.fao.org/faostat/en/#data/QCL) this appears as 'Low Income Food Deficit Countries (LIFDCs)'. Given that there is not always an ISO2/3 equivalent for each aggregate, one needs to pay attention to those cases. 

<!-- To add: perhaps turn this into a short checklist: identify aggregate rows; decide whether to keep, drop, or reconstruct them; document the choice; and compare constructed aggregates to a reference source when possible. -->

<!-- What if we want to aggregate multiple rows. I've removed before the EU as an aggregate, but let's assume that we'd want to construct it from scratch: -->
<!-- ```{r} -->
<!-- eu_df <- country_panel |>  -->
<!--   mutate( -->
<!--     iso2 = if_else(iso2 %in% eu_iso2, "EU", iso2), -->
<!--     country_name = if_else(iso2 == "EU", "European Union", country_name) -->
<!--   ) |> -->
<!--   summarise( -->
<!--     gdp_b = sum(gdp_b, na.rm = TRUE), -->
<!--     pop_m = sum(pop_m, na.rm = TRUE), -->
<!--     .by = c(iso2, country_name, year) -->
<!--   ) -->
<!-- ``` -->

<!-- # eu_comparison <- eu_df |>  -->
<!-- #   filter(iso2 == "EU") |>  -->
<!-- #   select(year, eu_constructed_gdp_b = gdp_b, eu_constructed_pop_m = pop_m) |> -->
<!-- #   left_join( -->
<!-- #     country_panel_all |>  -->
<!-- #       filter(iso3 == "EUU") |>  -->
<!-- #       select(year, eu_wdi_gdp_b = gdp_b, eu_wdi_pop_m = pop_m), -->
<!-- #     by = "year" -->
<!-- #   ) |> -->
<!-- #   mutate( -->
<!-- #     gdp_diff_b = round(eu_constructed_gdp_b - eu_wdi_gdp_b, 2), -->
<!-- #     pop_diff_m = eu_constructed_pop_m - eu_wdi_pop_m -->
<!-- #   ) -->
<!-- #  -->
<!-- # eu_comparison -->

<!-- Importantly for cases like this, ideally we'd want to verify that the values make sense compared to some reference data. For instance, I would try to find data on the aggregate (EU, in this case) from another source - IMF, Eurostat, etc, asking an AI whether it looks plausible etc.  -->

# Country Level Analysis

A lot of the analysis I have done was at the country level, or used countries as parent units. Thus, most of this text is going to use countries as example units, but here are some tips on dealing with country level data:

## Country Identifiers
Country names are useful for plots and tables, but they are often inconsistent across datasets. For example, "Ivory Coast" and "Cote d'Ivoire" may refer to the same country. For joins, it is usually better to rely on ISO codes, while keeping the names as labels.

Even country codes need a quick check. For example:
1) South Sudan's ISO3 is usually "SSD", but some datasets may use "SDS"
2) Namibia's ISO2 is "NA", which can be confused with a missing value if the data is read carelessly

 ISO codes are usually better than names, but still require some checks, given aggregates, disputed areas, territories, historical countries, and source-specific codes.

Normally the WDI includes data on aggregates (if you download directly with `WDI::WDI()`, use `extra = TRUE`). For illustration, however, I downloaded population and GDP from the WDI website as CSV files: https://data.worldbank.org/indicator/SP.POP.TOTL and https://data.worldbank.org/indicator/NY.GDP.MKTP.KD.

The construction code is not the important part here. The useful habit is to build the panel, then immediately inspect which country codes did not behave like ordinary countries.

```{r build-country-panel}
pop <- read_csv(here("data", "API_SP.POP.TOTL_DS2_en_csv_v2_33112.csv"), skip = 4) |>
  pivot_longer(
    cols = matches("^\\d{4}$"),
    names_to = "year",
    values_to = "pop"
  ) |>
  transmute(
    iso3 = `Country Code`,
    country_name = `Country Name`,
    year = as.integer(year),
    pop
  )

gdp <- read_csv(here("data", "API_NY.GDP.MKTP.KD_DS2_en_csv_v2_34247.csv"), skip = 4) |>
  pivot_longer(
    cols = matches("^\\d{4}$"),
    names_to = "year",
    values_to = "gdp"
  ) |>
  transmute(
    iso3 = `Country Code`,
    country_name = `Country Name`,
    year = as.integer(year),
    gdp
  )

country_panel_all <- full_join(pop, gdp, join_by(iso3, country_name, year)) |>
  transmute(
    iso2 = countrycode::countrycode(iso3, "iso3c", "iso2c", warn = FALSE),
    iso3,
    country_name,
    year,
    gdp_b = round(gdp / 1e9, 5),
    pop_m = round(pop / 1e6, 3)
  ) |>
  filter(year >= 2000) |>
  arrange(iso2, year)
```

Here is the resulting panel:

```{r inspect-country-panel}
head(country_panel_all)
```

This WDI file also includes rows that are not countries, such as regions or income groups. One quick diagnostic is to inspect the ISO3 codes that failed to convert to ISO2. Most failed conversions are aggregates, but this check also catches real places that need manual handling, like Kosovo.

```{r list-unmatched-countries}
unmatched_iso3 <- country_panel_all |> 
  filter(is.na(iso2)) |> 
  distinct(iso3, country_name) |> 
  arrange(country_name)

tibble::tibble(n_unmatched_iso3 = nrow(unmatched_iso3))
```

Rather than reading the whole unmatched list, I would inspect a few rows around the cases I care about:

```{r inspect-unmatched-countries}
kosovo_row <- which(unmatched_iso3$iso3 == "XKX")

unmatched_iso3 |> 
  slice(max(1, kosovo_row - 2):min(n(), kosovo_row + 2))
```

Thus, I'd manually add an ISO2 code for Kosovo. I also keep the EU as a custom aggregate because I want to use it later.

```{r fix-country-codes}
country_panel <- country_panel_all |> 
  mutate(
    iso2 = case_when(
      iso3 == "XKX" ~ "XK",
      iso3 == "EUU" ~ "EU", 
      TRUE ~ iso2
      )
  ) |> 
  filter(!is.na(iso2))
```

## Geographies Consistency Over Time

Various geographical units of observations do not have consistent boundaries over time. Countries change their borders, gain independence or split; admin units sometimes change have changed their borders. Different datasets treat those changes differently - perhaps a value in a given year refers to the current border or the old one. 
When working with country-level data, for example, pay attention to countries whose borders have changed in the period of analysis (or recently!). 'Sudan' might reflect the current state of Sudan, or also the values of South Sudan. Montenegro might be part of Yugoslavia, Serbia and Montenegro, or Montenegro; the same is true for Kosovo (despite not being mentioned in the name of 'Serbia and Montenegro'), although Kosovo might not even exist in certain datasets. Taiwan might be referred to in different ways, depending on the dataset.

<!-- To add: perhaps make this a checklist too: which boundaries does the dataset use; are units historical or current; were any units split/merged during the period; are successor states treated consistently; and do plots around the transition years reveal suspicious jumps. -->

This, for example, looks rather fine - there are no obvious spikes in the data around the year of independence:
```{r plot-sudan-south-sudan}
country_panel |>
  filter(iso2 %in% c("SD", "SS"), year %in% 2006:2014) |>
  ggplot() + 
  geom_line(aes(x = year, y = pop_m, group = country_name, color = country_name))

```

# Missing Values
Assume that we have a dataset with some missing values:
```{r make-na-example}
na_df <- country_panel |>
  filter(
    iso2 %in% c("CU", "ER", "BO"), 
    year %in% 2023:2025
    ) |>
  select(year, country_name, gdp_b) |> 
  pivot_wider(
    names_from = country_name,
    values_from = c(gdp_b)
  )

na_df
```

Missing values can mean different things - the value was not collected, it was not applicable, it was suppressed, it failed to merge, or it is genuinely unknown. R will often show these as `NA`. `NaN` is the result of an undefined numeric operation, such as `0 / 0`. `NULL` usually means that the object is absent. 

If we want to aggregate that data, the main options are 1) na.rm = TRUE, which removes the NAs before calculating, such that sum(df$x, na.rm = TRUE) would still return a value. For sums this can look like treating NA as 0, and if all values are NA it returns 0, but for means it also changes the denominator; 2) na.rm = FALSE, which forces the aggregation to have NA if any of the values is NA 

Especially if we use a loop or across to aggregate over several variables or units, it can be a problem
Say that we want to sum GDP, in billions, for each of those three countries in this period.  
```{r compare-na-sums}
na_df_long <- na_df |> 
  pivot_longer(
    -year,
    names_to = "country_name",
    values_to = "gdp_b"
  )

na_rm_options <- c("na.rm = TRUE", "na.rm = FALSE", "NA only if all missing")

bind_rows(
  na_df_long |> 
    summarise(
      method = "na.rm = TRUE",
      sum = sum(gdp_b, na.rm = TRUE),
      .by = country_name
    ),
  na_df_long |> 
    summarise(
      method = "na.rm = FALSE",
      sum = sum(gdp_b, na.rm = FALSE),
      .by = country_name
    ),
  na_df_long |> 
    summarise(
      method = "NA only if all missing",
      sum = if (all(is.na(gdp_b))) NA_real_ else sum(gdp_b, na.rm = TRUE),
      .by = country_name
    )
  ) |> 
  arrange(country_name, match(method, na_rm_options)) |> 
  pivot_wider(
    names_from = method,
    values_from = sum
  ) |> 
  left_join(
    na_df_long |> 
      summarise(
        n_observed = sum(!is.na(gdp_b)),
        .by = country_name
      ),
    by = "country_name"
  )
```

so the table lets us see the three choices together: when we use na.rm = FALSE, a column that has values and NAs is always NA. But perhaps we are interested in preserving the NAs only in cases where all rows are NA. then, we need to modify the function. It is also often useful to show the amount of missingness itself. Otherwise, two columns can have the same sum while one of them is based on much less data.

A small summary like this is often more useful than just calculating the aggregate:

```{r missingness-summary-example}
na_df_long |>
  summarise(
    n_rows = n(),
    n_missing = sum(is.na(gdp_b)),
    n_observed = sum(!is.na(gdp_b)),
    share_missing = n_missing / n_rows,
    .by = country_name
  )
```

# Plotting

Plots are great. I mainly use them for two, somewhat different purposes - inspecting the data and presenting results. Graphs can highlight weird data spikes, missing or implausible values, certain units driving the results in a (much) clearer way than summary tables. 
When presenting results, we want the readers to understand the main message, ideally without working too hard.
To me, graphs in economic papers tend to be quite bad in serving the latter goal; they're often ugly, too busy and hard to understand (whether it's due to lack of information, to low priority given to visualization, to standardization of those types of graphs or anything else is a different question). That is sometimes necessary, especially for readers who want to inspect the details, but often the graph becomes less useful because of it.

There are much better guides than whatever I will write here. I really like [Saloni Dattani's guide](https://www.scientificdiscovery.dev/p/salonis-guide-to-data-visualization).

One should consider which type of graph would be useful. I think that line plots are suitable for most cases, especially for changes over time, especially with a small number of series or when some series need to be highlighted. Bar plots can work well for comparing categories, if there are not too many of them; while stacked bars (or areas) can be nice at times but can be hard to read unless the question is really about the total, the first category in the stack, or categories that undertook massive changes. Pie charts should be avoided for similar reasons (but much harder to read)

The [R Graph Gallery](https://r-graph-gallery.com/) is also a nice source for examples.

Some things to consider:

Axis titles, especially y axis titles, are often unnecessary, hard to read, and distracting. The same information can usually be conveyed using the main plot title. Very often, the x axis indicates time, which means that indicating the unit is unnecessary, though it still appears on many graphs in economics papers.

```{r plot-gdp-pc-raw}
eu27_countries <- c(
  "AT", "BE", "BG", "HR", "CY", "CZ", "DK", "EE", "FI", "FR",
  "DE", "GR", "HU", "IE", "IT", "LV", "LT", "LU", "MT", "NL",
  "PL", "PT", "RO", "SK", "SI", "ES", "SE"
)
highlight_countries <- c("IE", "LU", "SE")

plot_df <- country_panel |> 
  filter(
    iso2 %in% eu27_countries,
    year >= 2000,
    year <= 2024
  ) |> 
  mutate(
    gdp_pc = gdp_b * 1000 / pop_m,
    highlighted = if_else(iso2 %in% highlight_countries, iso2, "Other countries")
  )

plot_df |> 
  ggplot() + 
  geom_line(aes(x = year, y = gdp_pc, color = iso2)) +
  scale_x_continuous(breaks = seq(2000, 2024, 4), limits = c(2000, 2024)) +
  scale_y_continuous(limits = c(0, NA), labels = label_dollar())
```
Very minimal adjustments make it a lot cleaner:
```{r plot-gdp-pc-clean}
plot_df |> 
  ggplot() + 
  geom_line(aes(x = year, y = gdp_pc, color = iso2)) + 
  scale_x_continuous(breaks = seq(2000, 2024, 4), limits = c(2000, 2024)) +
  scale_y_continuous(limits = c(0, NA), labels = label_dollar()) +
  theme_minimal() + 
  theme(axis.title = element_blank()) + 
  labs(title = "GDP per capita across EU countries")
```

The legend can also often be avoided by placing the labels directly near the lines. Wouldn't always work (generally useful if there aren't many categories), but it's sometimes useful. The plot can also benefit from clearer colours (that are more colour-blind friendly). Say that we want to focus on Ireland's rise from Sweden's level of GDP per capita to nearly Luxembourg's, then these could be highlighted with the other EU countries having grey:

```{r plot-gdp-pc-highlighted}
plot_colors <- c(
  "IE" = "#0072B2",
  "LU" = "#D55E00",
  "SE" = "#009E73",
  "Other countries" = "#b8bec6"
)

line_labels <- plot_df |> 
  filter(iso2 %in% highlight_countries) |> 
  summarise(
    gdp_pc = gdp_pc[which.max(year)],
    year = max(year),
    label = country_name[which.max(year)],
    .by = iso2
  )

plot_df |> 
  ggplot(aes(x = year, y = gdp_pc, group = iso2)) + 
  geom_line(aes(color = highlighted), linewidth = 0.9, alpha = 0.85) + 
  geom_text(
    data = line_labels,
    aes(label = iso2, color = iso2),
    hjust = 0,
    nudge_x = 0.25,
    show.legend = FALSE
  ) +
  scale_color_manual(values = plot_colors) +
  scale_x_continuous(breaks = seq(2000, 2024, 4), limits = c(2000, 2025.5)) +
  scale_y_continuous(limits = c(0, NA), labels = label_dollar()) +
  theme_minimal() + 
  theme(
    axis.title = element_blank(),
    legend.position = "none",
    plot.margin = margin(5.5, 60, 5.5, 5.5)
  ) + 
  coord_cartesian(clip = "off") +
  labs(title = "Ireland spiked toward the EU's richest")
```

Finally, a more polished version could use background colours, cleaner axis, using the actual names as labels and so on:

```{r plot-gdp-pc-final}
plot_df |> 
  ggplot(aes(x = year, y = gdp_pc, group = iso2)) + 
  geom_line(aes(color = highlighted), linewidth = 0.9, alpha = 0.8) + 
  geom_text(
    data = line_labels,
    aes(label = label, color = iso2),
    hjust = 0,
    nudge_x = 0.25,
    fontface = "bold",
    show.legend = FALSE
  ) +
  scale_color_manual(values = plot_colors) +
  scale_x_continuous(breaks = seq(2000, 2024, 4), limits = c(2000, 2025.5)) +
  scale_y_continuous(limits = c(0, NA), labels = label_dollar()) +
  labs(
    title = "Ireland spiked toward the EU's richest",
    subtitle = "GDP per capita, constant 2015 US dollars",
    caption = "Source: The World Bank"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.title = element_blank(),
    legend.position = "none",
    plot.background = element_rect(fill = "#f7f8f3", color = NA),
    panel.background = element_rect(fill = "#f7f8f3", color = NA),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    plot.caption = element_text(color = "#6b7280"),
    plot.margin = margin(5.5, 70, 5.5, 5.5)
  ) +
  coord_cartesian(clip = "off")
```

Another useful AI trick is to ask it to follow a specific visualization guide rather than asking for a "better plot" in general. For example, one could ask for suggestions inspired by the *Financial Times* visual vocabulary, the Urban Institute data visualization style guide, or the Datawrapper/Flourish style of direct labeling and restrained annotation.

# Spatial Analysis

Quite a lot of the analysis I have done is related to spatial data, so I include a few practical notes here.

There are two main types of spatial datasets. Vector data includes discrete geometries, such as points, lines, and polygons. Raster data stores values in grid cells. 

Pay attention to CRS and projections. Areas, distances, buffers, and centroids can be misleading if the data are in longitude/latitude or if layers use different CRS, so check `st_crs()`/`crs()` and use `st_transform` when needed to a projection that makes sense for the operation.

Note that vector data is often saved as shapefiles. Shapefiles are split across several files (`.shp`, `.shx`, `.dbf`, etc.), have short column-name limits, do not handle mixed geometry types well, and are not especially efficient. **`gpkg` is often preferable**: it is one file, handles longer column names, can store multiple layers, and is usually easier to move around.

**For vector data, I would usually use `sf`** in R. `sf::read_sf()` and `sf::write_sf()` are usually convenient wrappers around `st_read()` and `st_write()`, so I would use them most of the time (especially when reading a file with one obvious layer). 

**For raster data, I would usually use `terra`** in R. Raster data is often saved as `.tif`, especially for ordinary gridded layers. Another common format is `.nc` or NetCDF, which is often used for climate or other data with several dimensions. 

## Which Tool To Use

For me, there are mainly three types of tools that are useful in spatial analysis.

| Type | Examples | Main advantages | Main drawbacks |
|---|---|---|---|
| Programming languages | R with `sf` and `terra`; Python with `geopandas`, `rasterio`, `xarray`, etc. | Reproducible workflows; easy to combine spatial work with data cleaning, modeling, and plotting; good for local vector/raster operations | Local storage and memory can become a problem; worse for some visual validation |
| GIS software | QGIS, ArcGIS | Inspection, validation, manual exploration, and cartographic output | requires local storage; less convenient for some analysis workflows |
| Google Earth Engine | GEE Code Editor, GEE Python API | Hosted data catalogs; convenient raster and remote-sensing operations; interactive map checking; avoids many local storage problems | Less intuitive coding and debugging; quotas and server-side limits; less convenient for some reproducible project workflows |

**Programming languages** - can do most of the work well. Running things is rather fast for most cases and visualizations (e.g., `ggplot2::geom_sf()`) are decent. Mainly, they are very convenient when spatial work is a part of a larger data construction or analysis workflow. 

**GIS software** - ArcGIS is powerful but not free; QGIS is very good for visualizations and debugging or validating data (e.g., examine whether a spatial join worked properly or matched the wrong places). It is also possible to program there, but I find the programming less convenient than `sf` in R (decide for yourself, though; perhaps starting out with QGIS is easier), and it creates an additional layer of workflow (save the data, load it elsewhere, etc.). I personally like working with R though, especially with AI agents helping write and debug code. 

**GEE** - useful mainly, in my opinion, when the data already exists in the Earth Engine data catalog (https://developers.google.com/earth-engine/datasets) or the GEE community catalog (https://gee-community-catalog.org/). For example, using Hansen forest loss data or built-up data from GHSL is easier when one can avoid downloading and storing the data, and instead we'd run the relevant operations there. It can also be advantageous for raster or remote-sensing operations, such as filtering large image collections by date/cloud cover/location, combining multiple images, and using reducers over space or time. However, this is less of a reason to use GEE for every heavy spatial process now that there are quotas for noncommercial users. It is also good for validation and checking layers on a map. The main drawback of GEE, beyond quotas, is the coding experience: the Code Editor uses the Earth Engine JavaScript API, many objects are server-side, and errors can be opaque. GEE now has a Gemini assistant in the Code Editor (which I haven't tried yet), which requires a Gemini API key, so it is not necessarily unlimited/free. In terms of workflows, it can be annoying, although it's possible to use the Python API to keep all the code in a single project with other R/Python/Stata codes.

When unsure which tool to use, AIs would probably be useful to help you judge, and now the costs of writing a sample code in R and examining its efficiency are so low that it's probably worth starting from it.

### Recommendations For Working With GEE

If there is one set of units that will be used across multiple scripts, I would usually create a master feature collection and call it at the start of each script, for example `var worldUnits = ee.FeatureCollection('projects/ee-your_project/assets/worldUnits');`. Then, we can have a master script that loads the unit file and then calls separate modules.

I would pay attention to boundary datasets that AIs tend to suggest, such as `FAO/GAUL/2015/level0`. They can be useful, but if we later group by ISO3 or compare to country-level data, some territories or administrative units may need to be merged first. In those cases it can be better to have a separate GEE script that cleans the boundary file, matches names (e.g., to ISO3), merges territories that should belong to the same unit, and exports the result as a reusable GEE asset. Then the analysis scripts can call that asset.

## Spatial Matching

A unit may intersect several polygons, a point may fall just outside a boundary because the datasets use slightly different borders, or a raster cell may only partly overlap the polygon. In those cases we need to decide whether to keep all matches, use the largest overlap, weight by area or population, use the nearest feature, or drop ambiguous cases. Whatever choice is made, it is worth saving a few examples and plotting them.

I think it is useful to separate the cases, because the main mistakes differ:

| Match type | Common tools | Main question |
|---|---|---|
| point to polygon | `st_join(..., join = st_within)` | which unit contains each point? | 
| polygon to polygon | `st_intersection()` | which areas overlap, and by how much? |
| nearest vector match | `st_nearest_feature()`, `nngeo::st_nn()` | which feature is closest when there is no exact match? |
| raster to polygon | `terra::extract()` | how should grid-cell values be summarized within each polygon? |

For point-to-polygon matching, I would usually use `st_join(points, polygons, join = st_within)`, which attaches the polygon attributes to each point that falls inside a polygon. After this, I would check unmatched points.

For polygon-to-polygon matching, note that there are two `sf` functions with similar names: `st_intersects()` and `st_intersection()`. `st_intersection()` returns a new spatial dataset whose rows are the overlapping pieces. For example, if one country overlaps three ecological zones, the result can have three rows for that country, each with a smaller geometry and attributes from both the country layer and the ecological-zone layer. This is useful when we need to calculate overlap areas or area-weighted variables. `st_intersects()` only tells us which geometries touch or overlap, so it can be useful as a quicker diagnostic, or as a first filtering step before running a heavier `st_intersection()`.

For example, take two countries and overlay them with a 0.5 by 0.5 degree grid. Some grid cells cross the border, so after `st_intersection()` one original grid cell can become two country-grid pieces:

```{r plot-country-grid-overlap, message=FALSE, warning=FALSE}
overlap_countries <- rnaturalearth::ne_countries(
  scale = "small",
  returnclass = "sf"
) |>
  filter(iso_a2 %in% c("BE", "NL")) |>
  select(iso2 = iso_a2, country = name_long, geometry)

country_grid <- overlap_countries |>
  st_make_grid(cellsize = c(0.5, 0.5), square = TRUE) |>
  st_as_sf() |>
  mutate(grid_id = row_number()) |>
  st_filter(overlap_countries)

country_grid_overlap <- st_intersection(country_grid, overlap_countries)

country_grid_overlap$overlap_area_km2 <- as.numeric(
  st_area(st_transform(country_grid_overlap, 3035))
) / 1e6

ggplot() +
  geom_sf(data = overlap_countries, fill = "#f3f4f6", color = "#111827", linewidth = 0.35) +
  geom_sf(data = country_grid, fill = NA, color = "#9ca3af", linewidth = 0.25) +
  geom_sf(
    data = country_grid_overlap,
    aes(fill = country),
    alpha = 0.55,
    color = "white",
    linewidth = 0.12
  ) +
  coord_sf(datum = NA) +
  theme_void() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold")
  ) +
  labs(
    title = "Grid cells can overlap more than one country",
    subtitle = "Belgium and the Netherlands overlaid with a 0.5 degree grid",
    fill = NULL
  )
```

The table below shows grid cells that were split across both countries:

```{r inspect-country-grid-overlap}
country_grid_overlap |>
  st_drop_geometry() |>
  summarise(
    total_area_km2 = round(sum(overlap_area_km2), 1),
    share_in_belgium = round(
      sum(overlap_area_km2[country == "Belgium"]) / sum(overlap_area_km2),
      3
    ),
    .by = grid_id
  ) |>
  filter(share_in_belgium > 0, share_in_belgium < 1) |>
  slice_head(n = 6)
```

Nearest-feature matching can be useful for vector data when there is no exact match. `st_nearest_feature()` gives the single nearest feature, while `nngeo::st_nn()` is useful when we want several nearest neighbors, or when we only want matches within a given distance. For example, we might match a mine to the nearest road, but only if the road is within 10 km.

For raster-to-polygon matching, the question is usually how to summarize cells inside each polygon. For example, if we have a population raster and administrative units, we may want total population in each unit. If we have a temperature raster, we may want the mean. 
Pay attention to partial overlaps, as a cell may sit partly inside a polygon and partly outside it. In `terra::extract()`, the default does not return cell fractions; it extracts the selected raster cells and applies the function we ask for, such as `mean` or `sum`. Setting `weights = TRUE` returns approximate fractions of cells covered by the polygon, while `exact = TRUE` is more precise but slower. Those fractions are not always the right choice, but they can be useful. For example, for population we might sum `population * fraction`, while for temperature we might calculate a weighted mean.

## Creating Visualizations

For maps in R, `ggplot2` with `geom_sf()` is usually enough. A choropleth map usually needs two inputs: boundaries and a variable to color each area.

The example below is useful because it first highlights a bug. If we join by ISO2, France is not matched because the `rnaturalearth` file uses `-99` as its ISO2 code. That is exactly the kind of problem maps can help catch quickly.
```{r map-eu-gdp-pc, message=FALSE}
eu_gdp_2024 <- country_panel |>
  filter(year == 2024, iso2 %in% eu27_countries) |>
  mutate(gdp_pc = gdp_b * 1000 / pop_m) |>
  select(iso2, iso3, gdp_pc)

europe_boundaries <- rnaturalearth::ne_countries(
  scale = "small",
  returnclass = "sf"
) |> 
  select(iso2 = iso_a2, iso3 = adm0_a3, continent, geometry) |>
  filter(continent == "Europe")

bad_iso2_map <- europe_boundaries |>
  mutate(in_eu = iso2 %in% eu27_countries) |>
  left_join(
    eu_gdp_2024 |> select(iso2, gdp_pc),
    join_by(iso2)
  )

bad_iso2_map |>
  filter(iso3 %in% c("FRA", "DEU", "ESP")) |>
  st_drop_geometry() |>
  select(iso2_boundary = iso2, iso3, in_eu, gdp_pc)

bad_iso2_map |>
  ggplot() +
  geom_sf(fill = "#d9d9d9", color = "white", linewidth = 0.12) +
  geom_sf(
    data = bad_iso2_map |> filter(in_eu),
    aes(fill = gdp_pc),
    color = "white",
    linewidth = 0.12
  ) +
  scale_fill_viridis_c(
    labels = label_number(scale = 1 / 1000, suffix = "k"),
    na.value = "#eeeeee",
    guide = guide_colorbar(
      title = NULL,
      barwidth = grid::unit(160, "pt"),
      barheight = grid::unit(8, "pt"),
      label.position = "bottom"
    )
  ) +
  coord_sf(xlim = c(-12, 35), ylim = c(34, 72), expand = FALSE) +
  theme_void() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    plot.caption = element_text(color = "#6b7280")
  ) +
  labs(
    title = "GDP per capita across EU countries",
    subtitle = "Reveals a bad join: France is not merged",
    caption = "Source: The World Bank"
  )
```

The fix is to join by ISO3. For the final display, I would also usually consider binning the values rather than using a continuous color scale, with the bins depending on the comparison we want to make. The first version below keeps a wide top category, which is useful for separating poorer, middle, rich, and very rich countries. The second version uses the same palette but compresses the top and gives more detail in the middle and bottom. 
```{r map-eu-gdp-pc-fixed, message=FALSE}
eu_gdp_map <- europe_boundaries |>
  mutate(in_eu = iso3 %in% eu_gdp_2024$iso3) |>
  left_join(
    eu_gdp_2024 |> select(iso3, gdp_pc),
    join_by(iso3)
  ) |>
  mutate(
    gdp_pc_group_wide = cut(
      gdp_pc,
      breaks = c(0, 20000, 35000, 50000, 70000, Inf),
      labels = c("<$20k", "$20k-$35k", "$35k-$50k", "$50k-$70k", "$70k+"),
      right = FALSE
    ),
    gdp_pc_group_middle = cut(
      gdp_pc,
      breaks = c(0, 10000, 15000, 20000, 30000, 40000, Inf),
      labels = c("<$10k", "$10k-$15k", "$15k-$20k", "$20k-$30k", "$30k-$40k", "$40k+"),
      right = FALSE
    )
  )

eu_gdp_map |>
  ggplot() +
  geom_sf(fill = "#d9d9d9", color = "white", linewidth = 0.12) +
  geom_sf(
    data = eu_gdp_map |> filter(in_eu),
    aes(fill = gdp_pc_group_wide),
    color = "white",
    linewidth = 0.12
  ) +
  scale_fill_manual(
    values = c(
      "<$20k" = "#edf8fb",
      "$20k-$35k" = "#b2e2e2",
      "$35k-$50k" = "#66c2a4",
      "$50k-$70k" = "#2ca25f",
      "$70k+" = "#006d2c"
    ),
    na.value = "#eeeeee",
    name = NULL
  ) +
  coord_sf(xlim = c(-12, 35), ylim = c(34, 72), expand = FALSE) +
  theme_void() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    plot.caption = element_text(color = "#6b7280")
  ) +
  labs(
    title = "GDP per capita across EU countries",
    subtitle = "Constant 2015 US dollars, not PPP-adjusted",
    caption = "Source: The World Bank"
  )

eu_gdp_map |>
  ggplot() +
  geom_sf(fill = "#d9d9d9", color = "white", linewidth = 0.12) +
  geom_sf(
    data = eu_gdp_map |> filter(in_eu),
    aes(fill = gdp_pc_group_middle),
    color = "white",
    linewidth = 0.12
  ) +
  scale_fill_manual(
    values = c(
      "<$10k" = "#edf8fb",
      "$10k-$15k" = "#ccece6",
      "$15k-$20k" = "#99d8c9",
      "$20k-$30k" = "#66c2a4",
      "$30k-$40k" = "#2ca25f",
      "$40k+" = "#006d2c"
    ),
    na.value = "#eeeeee",
    name = NULL
  ) +
  coord_sf(xlim = c(-12, 35), ylim = c(34, 72), expand = FALSE) +
  theme_void() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    plot.caption = element_text(color = "#6b7280")
  ) +
  labs(
    title = "GDP per capita across EU countries",
    subtitle = "Same palette, narrower bins",
    caption = "Source: The World Bank"
  )
```

## Finding Representative Points Of Polygons
Sometimes we want to use single points in a given unit, rather than the full polygon. That's especially the case if the analysis is very heavy (e.g., calculating road path from unit x to y), or we are interested in the role of specific areas within the unit.

How to select the point? These are some options, and the choice depends on what we want to capture:

| Point choice | Useful when | Main concern |
|---|---|---|
| centroid | geometry itself is the object of interest | can fall outside the polygon or in an unimportant area |
| `st_point_on_surface()` | point must be inside the polygon | can still land in a substantively odd place |
| population-weighted point | people are the object of interest | may fall between population centers or outside the unit |
| highest-population cell / main city | one central populated place is wanted | depends heavily on resolution of population data |

```{r plot-centroid-surface-points, echo=FALSE}
normal_polygon <- st_polygon(list(
  matrix(
    c(
      0.2, 0.1,
      4.1, 0.0,
      4.4, 2.2,
      3.6, 4.0,
      1.3, 4.3,
      0.0, 2.5,
      0.2, 0.1
    ),
    ncol = 2,
    byrow = TRUE
  )
))

hole_polygon <- st_polygon(list(
  matrix(c(0, 0, 4, 0, 4, 4, 0, 4, 0, 0), ncol = 2, byrow = TRUE),
  matrix(c(1.2, 1.2, 1.2, 2.8, 2.8, 2.8, 2.8, 1.2, 1.2, 1.2), ncol = 2, byrow = TRUE)
))

schematic_units <- st_sf(
  case = c("compact polygon", "polygon with hole"),
  geometry = st_sfc(normal_polygon, hole_polygon)
)

centroids <- schematic_units |>
  st_centroid() |>
  mutate(point_type = "centroid")

surface_points <- schematic_units |>
  st_point_on_surface() |>
  mutate(point_type = "point on surface")

bind_rows(centroids, surface_points) |>
  ggplot() +
  geom_sf(data = schematic_units, fill = "#e5e7eb", color = "#4b5563") +
  geom_sf(aes(color = point_type), size = 2.8) +
  facet_wrap(~case) +
  scale_color_manual(values = c("centroid" = "#D55E00", "point on surface" = "#0072B2")) +
  coord_sf(datum = NA) +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.title = element_blank()
  )
```

```{r plot-population-centroid-points, echo=FALSE, fig.width=6, fig.height=4}
population_unit <- st_sf(
  geometry = st_sfc(normal_polygon)
)

mock_population_grid <- st_sf(
  geometry = st_make_grid(population_unit, n = c(10, 10))
) |>
  st_intersection(population_unit)

grid_centers <- st_coordinates(st_centroid(mock_population_grid))

mock_population_grid <- mock_population_grid |>
  mutate(
    x_coord = grid_centers[, 1],
    y_coord = grid_centers[, 2],
    population = round(
      15 +
        950 * exp(-((x_coord - 1.0)^2 + (y_coord - 3.0)^2) / 0.18) +
        240 * exp(-((x_coord - 3.2)^2 + (y_coord - 1.0)^2) / 0.14) +
        180 * exp(-((x_coord - 3.4)^2 + (y_coord - 3.1)^2) / 0.16) +
        130 * exp(-((x_coord - 2.0)^2 + (y_coord - 1.9)^2) / 0.12)
    )
  )

population_weighted_point <- mock_population_grid |>
  st_drop_geometry() |>
  summarise(
    x_coord = weighted.mean(x_coord, population),
    y_coord = weighted.mean(y_coord, population)
  ) |>
  st_as_sf(coords = c("x_coord", "y_coord")) |>
  mutate(point_type = "population-weighted centroid")

highest_population_point <- mock_population_grid |>
  slice_max(population, n = 1, with_ties = FALSE) |>
  st_centroid() |>
  mutate(point_type = "highest population cell")

population_centroid <- population_unit |>
  st_centroid() |>
  mutate(point_type = "centroid")

population_surface_point <- population_unit |>
  st_point_on_surface() |>
  mutate(point_type = "point on surface")

population_summary_points <- bind_rows(
  population_centroid,
  population_surface_point,
  population_weighted_point,
  highest_population_point
)

population_summary_labels <- population_summary_points |>
  mutate(
    point_x = st_coordinates(geometry)[, 1],
    point_y = st_coordinates(geometry)[, 2],
    label_x = case_when(
      point_type == "highest population cell" ~ point_x - 0.16,
      point_type == "population-weighted centroid" ~ point_x + 0.18,
      point_type == "centroid" ~ point_x + 0.18,
      TRUE ~ point_x - 0.16
    ),
    label_y = case_when(
      point_type == "highest population cell" ~ point_y + 0.20,
      point_type == "population-weighted centroid" ~ point_y - 0.18,
      point_type == "centroid" ~ point_y + 0.18,
      TRUE ~ point_y - 0.20
    )
  ) |>
  st_drop_geometry()

ggplot() +
  geom_sf(
    data = mock_population_grid,
    aes(fill = population),
    color = "white",
    linewidth = 0.2
  ) +
  geom_sf(data = population_unit, fill = NA, color = "#4b5563", linewidth = 0.5) +
  geom_sf(
    data = population_summary_points,
    aes(color = point_type),
    size = 3
  ) +
  geom_label(
    data = population_summary_labels,
    aes(x = label_x, y = label_y, label = point_type),
    size = 3.1,
    linewidth = 0.15,
    label.padding = unit(0.08, "lines")
  ) +
  scale_fill_gradient(low = "white", high = "#b91c1c") +
  scale_color_manual(
    values = c(
      "centroid" = "#D55E00",
      "point on surface" = "#0072B2",
      "population-weighted centroid" = "#009E73",
      "highest population cell" = "#00BFC4"
    )
  ) +
  coord_sf(datum = NA, expand = TRUE) +
  theme_minimal(base_size = 12) +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  )
```

### Finding A 'Representative' Category
there are various spatial datasets that might involve finding some balance between various categories. mines datasets, for example, sometimes have each individual mine as a row, or would have a value of 'suitability for a given commodity' per cell. depends on what we want to have, we might want to aggregate the data per unit. should we sum the number of distinct mines? use the commodity with the highest value of suitability? count whether any mine of each type exists? There is no obvious answer here. The right choice depends on whether the variable is meant to capture abundance, dominance, exposure, or just the presence of something.

# Analysis

## Shift Share

to be added

# Debugging And Validating Results

A lot of mistakes are much easier to catch immediately after the step that created them. I would usually run some checks after important joins, aggregations, spatial matches, and output construction, rather than only looking at the final regression table or figure.

Useful checks:

- row and column counts before and after
- duplicate keys before joins
- unmatched observations after joins
- row multiplication after joins, meaning that the join created more rows than expected because one or both datasets had repeated keys
- missingness summaries for new variables - similarly with zeros (where zeros indicate a potential problem)
- impossible or implausible values
- extreme values
- edge cases and specific examples
- graphs, especially over time or on maps
- compare values, when possible, to an external source (e.g., after cleaning WB data and merging countries, if unsure - check IMF or publications online and see if the values are reasonably close)
- inspect some random rows

Before sending results, I would usually rerun the relevant script from a clean session and check that the expected outputs were recreated. If something looks odd but is probably fine, it is still worth mentioning it briefly rather than letting the next person rediscover it.

For example, after adding GDP to a country-region file, I might run something like this before trusting the join:

```{r join-validation-example}
join_check <- country_regions |>
  left_join(
    country_gdp |> select(-country_name),
    join_by(iso3)
  )

tibble::tibble(
  rows_before = nrow(country_regions),
  rows_after = nrow(join_check),
  unmatched_iso3 = nrow(anti_join(country_regions, country_gdp, join_by(iso3))),
  missing_gdp = sum(is.na(join_check$gdp_b))
)

anti_join(country_regions, country_gdp, join_by(iso3))
```

# Constructing Workflows

I prefer working inside an R project, which makes relative paths more stable. Functions like `here::here()` are useful for referring files from the project root.

I like organising both code and data in raw, intermediate/build, and final/output folders. raw data should usually be kept as close as possible to the original files, intermediate data should be the products of construction steps, and final/output data should be the version that is actually used for analysis, tables, or figures. This makes it easier to know what can be recreated and what should not be manually edited.

For code, I usually prefer numbered scripts or folders when there is a logical order, for example `01_import`, `02_clean`, `03_construct`, `04_analysis`. 

If some construction steps are slow, it can be worth saving intermediate outputs and reusing them, but I would give them clear names and avoid keeping several unexplained versions of the same file.

Version control (usually git, and often GitHub) is extremely useful. In projects with many authors, it can help deal with conflicts and make it clearer who changed what; but even for a project that only you work on, it is recommended, for tracking changes over time and reverting them more easily. I would usually make small commits with messages that explain the change, and use branches or pull requests when the project has several people or when a change is risky. For (especially heavy) data files, outputs that are repeatedly generated with different names (e.g., results_\{date\}) or private files, should go in `.gitignore` and not be tracked.

For reproducibility, I would keep the seed discussion near the code that actually uses randomness. Fixed seeds (e.g., `set.seed(123)`) are deterministic - see [this blog post](https://blog.genesmindsmachines.com/p/if-your-random-seed-is-42-i-will). One option is to generate a seed once, record it, and then keep using that same seed throughout the project.

Before sending results, I would rerun the relevant scripts from a clean session and check the main outputs. This usually means row counts, duplicate keys, unmatched observations, summary statistics, missing values, and zeros in places where zeros may indicate a construction problem.

## Handoff Notes

When handing work to someone else, I would usually include a short note rather than only sending the files. Useful things to mention:

- what changed
- which scripts or files matter
- where the main outputs are
- which checks were run
- assumptions or manual decisions
- strange cases that are probably fine
- things that are unfinished or need judgment

## Definitions And Units

Pay attention to the definition and the units. `gdp` could mean nominal GDP in current local currency, nominal GDP in current US dollars, real GDP in constant 2015 US dollars, or PPP-adjusted GDP, while `gdp_pc` also depends on the population denominator. Different definitions have different meanings (and often, differences in the number of missing values).

For example, using the WDI variables above, `gdp_b` is GDP in billions of constant 2015 US dollars, `pop_m` is population in millions, and `gdp_pc = gdp_b * 1000 / pop_m` gives GDP per person in dollars. That is not the same as GDP per capita in current dollars, local currency, or PPP-adjusted terms, and those differences can matter a lot for comparisons across countries and over time.

Even with the existing data, using total GDP or GDP per capita can imply a very different analysis. Total GDP is partly a size variable, while GDP per capita is closer to an income/productivity measure:

```{r compare-gdp-definitions}
country_panel |>
  filter(year == 2024) |>
  mutate(
    gdp_pc = gdp_b * 1000 / pop_m,
    log_gdp_pc = log(gdp_pc)
  ) |>
  summarise(
    cor_gdp_pop = cor(gdp_b, pop_m, use = "complete.obs"),
    cor_gdp_pc_pop = cor(gdp_pc, pop_m, use = "complete.obs"),
    cor_log_gdp_pc_pop = cor(log_gdp_pc, pop_m, use = "complete.obs")
  )
```

If using `log()`, check the values and units first. Logging GDP in dollars is not the same as logging GDP in millions, and logging very small values can produce numbers that are technically correct but not very meaningful. 

Labels and categories also need some care. Value labels can be useful for understanding survey or administrative data, but they can disappear, change, or conflict after transformations; and factors should be checked before recoding, joining, or modelling.

## Assumptions And Judgment Calls

Some choices are not purely technical, and it is better to document them explicitly. For example: dropping aggregate regions, manually coding Kosovo, using the largest spatial overlap, treating all-missing sums as `NA`, using current boundaries for historical data, or choosing one representative point for a polygon. Even a short note is useful if someone later needs to understand why the data look the way they do.

# AI Use

This would probably change quite a bit soon, but the setup that I like now for working with AI agents is to separate reusable preferences from project-specific context. One possible structure is:

```text
ai_instructions/
  general/
    coding_style.md
    data_checks.md
    writing_style.md
  project/
    context.md
    pipeline.md
    datasets.md
```

The general files can include coding conventions, for example using `|>` rather than `%>%`, and recurring data-construction checks, such as looking for missing values after merges or joins that multiply rows. The project files can describe the question, the main unit of observation, the code structure, and the intended order of the data pipeline. This does use context, so I would not overdo it for small tasks, but for longer projects it can save repeated explanations and reduce mistakes. If there are project-specific coding conventions, I would usually add them as short additions to the general convention rather than copying the whole thing again.

There are also many good skills online, so there is probably no need to create many new ones except for specific cases. I like using things such as humanizer and deep research, and it can be useful to take an existing skill and adjust it to your preferences. For example, my coding convention skill is edited from https://rworks.dev/posts/claude-skills-for-r-users/.

For tasks that are not obvious, I prefer to start in plan mode or ask the AI to restate the task before implementing. Sometimes I ask it to write the prompt back to me, then I edit the prompt and use it in a new chat. This can help avoid very long chats and makes unclear assumptions more visible. It is also useful to ask the AI to interview you when there are missing choices. For example:

```text
Bad: fix this merge

Better: use the project context and data-check instructions, fix the country-year merge, and report duplicate keys, unmatched rows, and any row-count changes
```

### AI And Validation

AI is useful for writing code, explaining unfamiliar packages, and finding possible bugs, but plausible code can still do the wrong thing. I would ask it to add messages and checks, summarize what changed, inspect examples rather than only summary statistics, and compare important constructed variables to an external source when possible.


# Specific Datasets

## WDPA (World Database on Protected Areas)

**Points** - The database includes protected areas (PAs) represented both as polygons and as points. For polygons, the boundaries of the PAs are available directly. Points are more ambiguous, as they can represent point-like features, or refer to larger protected areas for which the shape is unknown. The `REP_AREA` field can be used to approximate the size of these point records, but the point location is not necessarily the centroid of the area, and the shape of the area is unknown. Therefore, buffering points by their reported area may be useful as an approximation, but it is unlikely to recover the true shape of the PA.

**Overlaps** - When intersecting WDPA with other spatial data, for example to estimate the share of an area covered by PAs, need to account for overlapping of PAs. A given location can be covered by multiple PAs, so simply summing areas across records can overstate coverage. Thus, protected-area geometries may need to be unioned before calculating total covered area (perhaps by category, depending on the goal).

**Status year** - The time variable in WDPA is `STATUS_YR`, which records the last year of status change. This usually refers to the legal designation as a protected area. In checks against South African protected-area data from the late 1990s onward, most protected areas did not appear to change status after their initial designation during the observed period. This suggests that `STATUS_YR` can often be informative about designation timing, but it should still be used cautiously, especially for cases involving expansions, reclassification, proposed or established areas, and loss of status.

## IPUMS

# Suggestions For Stuff To Add / Remove

- add a short opening paragraph about what kinds of RA work this is based on
- add one concrete example of a result validation workflow
- add a small checklist at the end
- maybe remove or shorten any examples that are too project-specific once the structure is clearer
