---
title: "RA lessons"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: hide
    css: styles.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# Introduction

this wouldnt speak much about measurement issues in datasets

```{r load-packages}
pacman::p_load(
  dplyr, labelled, sjlabelled, tidyr, ggplot2, scales, sf,
  here, readr, dtplyr, duckdb, DBI
  )

options(scipen = 999)
```

# Working in R

## style and speed

I mainly used `dplyr`, due to a combination of preference for its syntax and the piped structure, and being used to it. 


```{r make-style-example}
example_df <- tibble::tibble(
  iso2 = c("US", "US", "CN", "CN"),
  gdp_t = c(25, 26, 17, 18)
)
```

- highly recommend to examine updates to the dplyr package (and in general, to packages that are used often). 
- the new pipe (|>) is nicer and cleaner than the old magrittr `%>%` pipe

see the difference between coding styles:

```{r old-pipe-style, results='hide', class.source='fold-show'}
example_df %>% 
  group_by(iso2) %>%
  summarise(
    avg_gdp = mean(gdp_t)
    ) %>%
  ungroup()
```

```{r native-pipe-style, results='hide', class.source='fold-show'}
example_df |> 
  summarise(
    avg_gdp = mean(gdp_t),
    .by = iso2
    ) 
```
Decide for yourself but I much prefer the latter.

`data.table` is faster, but for most processes the marginal improvement is rather small. 

For heavier processes, one can use `dtplyr`. For smaller processes, it seems better to use `dplyr` regularly, as the setup costs may take longer than the actual calculation.

```{r benchmark-dplyr-dtplyr}
set.seed(23072026)

speed_df <- tibble::tibble(
  iso2 = sample(sprintf("C%03d", 1:200), 3000000, replace = TRUE),
  year = sample(2000:2024, 3000000, replace = TRUE),
  gdp_t = runif(3000000, 1, 100),
  pop_m = runif(3000000, 1, 1500)
)

dplyr_elapsed <- system.time({
  dplyr_result <- speed_df |>
    mutate(
      gdp_pc = gdp_t * 1000000 / pop_m,
      large = gdp_t > 50
    ) |>
    filter(large, year >= 2010) |>
    summarise(
      avg_gdp_pc = mean(gdp_pc),
      total_gdp = sum(gdp_t),
      n = n(),
      .by = c(iso2, year)
      )
})[["elapsed"]]

cat("dplyr elapsed:", round(dplyr_elapsed, 3), "seconds\n")

dtplyr_elapsed <- system.time({
  dtplyr_result <- speed_df |>
    lazy_dt() |>
    mutate(
      gdp_pc = gdp_t * 1000000 / pop_m,
      large = gdp_t > 50
    ) |>
    filter(large, year >= 2010) |>
    summarise(
      avg_gdp_pc = mean(gdp_pc),
      total_gdp = sum(gdp_t),
      n = n(),
      .by = c(iso2, year)
      ) |>
    as_tibble()
})[["elapsed"]]

cat("dtplyr elapsed:", round(dtplyr_elapsed, 3), "seconds\n")
```

`duckdb` is another fast option, especially when the data is too large to comfortably keep in memory. It requires SQL syntax, however.

```{r benchmark-duckdb}
con <- dbConnect(duckdb())
duckdb_register(con, "speed_df", speed_df)

duckdb_elapsed <- system.time({
  duckdb_result <- dbGetQuery(
    con,
    "
    SELECT
      iso2,
      year,
      AVG(gdp_t * 1000000 / pop_m) AS avg_gdp_pc,
      SUM(gdp_t) AS total_gdp,
      COUNT(*) AS n
    FROM speed_df
    WHERE gdp_t > 50
      AND year >= 2010
    GROUP BY iso2, year
    "
  )
})[["elapsed"]]

cat("duckdb elapsed:", round(duckdb_elapsed, 3), "seconds\n")

dbDisconnect(con, shutdown = TRUE)
```

## functions with the same names 

Some R functions from different packages have the identical names. This is the source of some bugs and unexpected behaviour, that is sometimes quite hard to detect. 
The `lag` function is a nice example of that. Its base version, `stats::lag`, is quite problematic as it doesn't actually lag the variables. `dplyr::lag` (or `data.table::shift(type = "lag")`) are much preferable. However, not specifying the package for this function may cause the use of `stats::lag`, and copy the values rather than lag them.

```{r compare-lag-functions}
lag_example <- tibble::tibble(
  year = 2020:2022,
  gdp = c(100, 120, 140)
)

lag_example |>
  mutate(
    `dplyr::lag` = dplyr::lag(gdp),
    `stats::lag` = as.numeric(stats::lag(gdp))
  )
```

Similarly, if the code uses `select` without specifying that it should call `dplyr::select`, while `raster` is called, the code might break given the existence of `raster::select`. Here, at least, there wouldn't be a mysterious bug, unlike the `lag` case). 

Another example is `copy_labels`, which exists in both `sjlabelled` and `labelled`. The two versions do not use the same argument order, and they may also change values while copying labels, so it is worth writing the package name explicitly and checking the documentation.

# Data construction and cleaning

## Joins

- discuss key uniqueness before joining
- discuss conflicting columns after a join, e.g., `country_name.x` and `country_name.y`
- discuss adding suffixes or selecting away unnecessary columns before joining
- discuss what happens to the rows under `left_join`, `inner_join`, etc
- discuss one-to-one, many-to-one, and many-to-many joins
- briefly mention that `full_join` and `right_join` also exist, but are needed less often

Assume that we have those two datasets, and we want to add the GDP variable into the country-region data:

```{r simple-country-join-example}
country_gdp <- tibble::tibble(
  iso3 = c("USA", "CHN", "IND"),
  country_name = c("United States", "China", "India"),
  gdp_b = c(23000, 18000, 3500)
)

country_regions <- tibble::tibble(
  iso3 = c("USA", "CHN", "IND", "MEX"),
  country_name = c("United States of America", "China", "India", "Mexico"),
  region = c("North America", "East Asia", "South Asia", "Latin America")
)

country_gdp

country_regions

```

```{r}

joined_country_regions <- country_regions |>
  left_join(country_gdp, join_by(iso3))

joined_country_regions

country_regions |>
  inner_join(country_gdp, join_by(iso3))

country_regions |>
  left_join(country_gdp, join_by(country_name))

country_regions |>
  left_join(
    country_gdp,
    join_by(iso3),
    suffix = c("_regions", "_gdp")
  )

country_regions |>
  left_join(
    country_gdp |> 
      select(-country_name),
    join_by(iso3)
  )
```

Here, both `left_join` and `inner_join` can add GDP, but they do not keep the same rows: `left_join` keeps Mexico and leaves GDP as `NA`, while `inner_join` drops Mexico. `full_join` and `right_join` also exist, but I use them less often.

In this example, joining by country name is problematic because the United States is written differently across the two datasets. The `country_name` columns also conflict after joining by `iso3`, so one option is to use suffixes and another is to drop the extra `country_name` column before joining.

## Messages and quick checks

- discuss adding messages after important data construction steps
- discuss reporting how many units joined, did not join, or have missing values after a join
- discuss reporting how many units were aggregated, how many rows were collapsed, or how many groups have missing values after `summarise`

```{r join-messages, message=TRUE}
joined_country_regions <- country_regions |>
  left_join(
    country_gdp |> 
      select(-country_name),
    join_by(iso3)
  )

unmatched_country_regions <- country_regions |>
  filter(!iso3 %in% country_gdp$iso3)

message(
  n_distinct(country_regions$iso3) - nrow(unmatched_country_regions), " of ",
  n_distinct(country_regions$iso3), " country-region units matched to GDP."
)

message(
  nrow(unmatched_country_regions), " country-region units did not match to GDP."
)

message(
  sum(is.na(joined_country_regions$gdp_b)), " rows have missing GDP after the join."
)
```

## Aggregates

In some datasets there may be aggregates (e.g., "Middle East" in a country-level dataset). In some datasets there are variables that refer to those, but not always, and the names of the 'aggregate' countries might be different across different datasets (e.g., in two different datasets of FAOSTAT, there is in one dataset (https://www.fao.org/faostat/en/#data/FBSH) an aggregated region of 'Low Income Food Deficit Countries', but in another (https://www.fao.org/faostat/en/#data/QCL) this appears as 'Low Income Food Deficit Countries (LIFDCs)'. Given that there is not always an ISO2/3 equivalent for each aggregate, one needs to pay attention to those cases. 

<!-- What if we want to aggregate multiple rows. I've removed before the EU as an aggregate, but let's assume that we'd want to construct it from scratch: -->
<!-- ```{r} -->
<!-- eu_df <- country_panel |>  -->
<!--   mutate( -->
<!--     iso2 = if_else(iso2 %in% eu_iso2, "EU", iso2), -->
<!--     country_name = if_else(iso2 == "EU", "European Union", country_name) -->
<!--   ) |> -->
<!--   summarise( -->
<!--     gdp_b = sum(gdp_b, na.rm = TRUE), -->
<!--     pop_m = sum(pop_m, na.rm = TRUE), -->
<!--     .by = c(iso2, country_name, year) -->
<!--   ) -->
<!-- ``` -->

<!-- # eu_comparison <- eu_df |>  -->
<!-- #   filter(iso2 == "EU") |>  -->
<!-- #   select(year, eu_constructed_gdp_b = gdp_b, eu_constructed_pop_m = pop_m) |> -->
<!-- #   left_join( -->
<!-- #     country_panel_all |>  -->
<!-- #       filter(iso3 == "EUU") |>  -->
<!-- #       select(year, eu_wdi_gdp_b = gdp_b, eu_wdi_pop_m = pop_m), -->
<!-- #     by = "year" -->
<!-- #   ) |> -->
<!-- #   mutate( -->
<!-- #     gdp_diff_b = round(eu_constructed_gdp_b - eu_wdi_gdp_b, 2), -->
<!-- #     pop_diff_m = eu_constructed_pop_m - eu_wdi_pop_m -->
<!-- #   ) -->
<!-- #  -->
<!-- # eu_comparison -->

<!-- Importantly for cases like this, ideally we'd want to verify that the values make sense compared to some reference data. For instance, I would try to find data on the aggregate (EU, in this case) from another source - IMF, Eurostat, etc, asking an AI whether it looks plausible etc.  -->

# Country level analysis

A lot of the analysis I have done was at the country level, or used countries as parent units. Thus, most of this text is going to use countries as example units, but here are some tips on dealing with country level data:

## Country identifiers
Country names are useful for plots and tables, but they are often inconsistent across datasets. For example, "Ivory Coast" and "Cote d'Ivoire" may refer to the same country. For joins, it is usually better to rely on ISO codes, while keeping the names as labels.

Even country codes need a quick check. For example:
1) South Sudan's ISO3 is usually "SSD", but some datasets may use "SDS"
2) Namibia's ISO2 is "NA", which can be confused with a missing value if the data is read carelessly

Normally the WDI includes data on aggregates (if you download directly with 'WDI::WDI(), use extra = TRUE')). For illustration, however, I have downloaded the data from the WDI website as csv. https://data.worldbank.org/indicator/SP.POP.TOTL and https://data.worldbank.org/indicator/NY.GDP.MKTP.KD

```{r build-country-panel}
pop <- read_csv(here("data", "API_SP.POP.TOTL_DS2_en_csv_v2_33112.csv"), skip = 4) |>
  pivot_longer(
    cols = matches("^\\d{4}$"),
    names_to = "year",
    values_to = "pop"
  ) |>
  transmute(
    iso3 = `Country Code`,
    country_name = `Country Name`,
    year = as.integer(year),
    pop
  )

gdp <- read_csv(here("data", "API_NY.GDP.MKTP.KD_DS2_en_csv_v2_34247.csv"), skip = 4) |>
  pivot_longer(
    cols = matches("^\\d{4}$"),
    names_to = "year",
    values_to = "gdp"
  ) |>
  transmute(
    iso3 = `Country Code`,
    country_name = `Country Name`,
    year = as.integer(year),
    gdp
  )

country_panel_all <- full_join(pop, gdp, by = c("iso3", "country_name", "year")) |>
  transmute(
    iso2 = countrycode::countrycode(iso3, "iso3c", "iso2c", warn = FALSE),
    iso3,
    country_name,
    year,
    gdp_b = round(gdp / 1e9, 5),
    pop_m = round(pop / 1e6, 3)
  ) |>
  filter(year >= 2000) |>
  arrange(iso2, year)

head(country_panel_all)
```

This WDI file also includes rows that are not countries, such as regions or income groups. One quick diagnostic is to inspect the ISO3 codes that failed to convert to ISO2. Most failed conversions are aggregates, but this check also catches real places that need manual handling, like Kosovo.

```{r list-unmatched-countries}
unmatched_iso3 <- country_panel_all |> 
  filter(is.na(iso2)) |> 
  distinct(iso3, country_name) |> 
  arrange(country_name)

tibble::tibble(n_unmatched_iso3 = nrow(unmatched_iso3))

kosovo_row <- which(unmatched_iso3$iso3 == "XKX")

unmatched_iso3 |> 
  slice(max(1, kosovo_row - 2):min(n(), kosovo_row + 2))
```

Thus, I'd manually add an ISO2 code for Kosovo. I also keep the EU as a custom aggregate because I want to use it later.

```{r fix-country-codes}
country_panel <- country_panel_all |> 
  mutate(
    iso2 = case_when(
      iso3 == "XKX" ~ "XK",
      iso3 == "EUU" ~ "EU", 
      TRUE ~ iso2
      )
  ) |> 
  filter(!is.na(iso2))
```

## Geographies consistency over time

Various geographical units of observations do not have consistent boundaries over time. Countries change their borders, gain independence or split; admin units sometimes change have changed their borders. Different datasets treat those changes differently - perhaps a value in a given year refers to the current border or the old one. 
When working with country-level data, for example, pay attention to countries whose borders have changed in the period of analysis (or recently!). 'Sudan' might reflect the current state of Sudan, or also the values of South Sudan. Montenegro might be part of Yugoslavia, Serbia and Montenegro, or Montenegro; the same is true for Kosovo (despite not being mentioned in the name of 'Serbia and Montenegro'), although Kosovo might not even exist in certain datasets. Taiwan might be referred to in different ways, depending on the dataset.

This, for example, looks rather fine - there are no obvious spikes in the data around the year of independence:
```{r plot-sudan-south-sudan}
country_panel |>
  filter(iso2 %in% c("SD", "SS"), year %in% 2006:2014) |>
  ggplot() + 
  geom_line(aes(x = year, y = pop_m, group = country_name, color = country_name))

```

# Missing values
Assume that we have a dataset with some missing values:
```{r make-na-example}
na_df <- country_panel |>
  filter(
    iso2 %in% c("CU", "ER", "BO"), 
    year %in% 2023:2025
    ) |>
  select(year, country_name, gdp_b) |> 
  pivot_wider(
    names_from = country_name,
    values_from = c(gdp_b)
  )

na_df
```
We want to aggregate that data. The main options are 1) na.rm = TRUE, which removes the NAs before calculating, such that sum(df$x, na.rm = TRUE) would still return a value. For sums this can look like treating NA as 0, and if all values are NA it returns 0, but for means it also changes the denominator; 2) na.rm = FALSE, which forces the aggregation to have NA if any of the values is NA 

Especially if we use a loop or across to aggregate over several variables or units, it can be a problem
Say that we want to sum GDP, in billions, for each of those three countries in this period.  
```{r compare-na-sums}
na_df_long <- na_df |> 
  pivot_longer(
    -year,
    names_to = "country_name",
    values_to = "gdp_b"
  )

na_rm_options <- c("na.rm = TRUE", "na.rm = FALSE", "NA only if all missing")

bind_rows(
  na_df_long |> 
    summarise(
      method = "na.rm = TRUE",
      sum = sum(gdp_b, na.rm = TRUE),
      .by = country_name
    ),
  na_df_long |> 
    summarise(
      method = "na.rm = FALSE",
      sum = sum(gdp_b, na.rm = FALSE),
      .by = country_name
    ),
  na_df_long |> 
    summarise(
      method = "NA only if all missing",
      sum = if (all(is.na(gdp_b))) NA_real_ else sum(gdp_b, na.rm = TRUE),
      .by = country_name
    )
  ) |> 
  arrange(country_name, match(method, na_rm_options)) |> 
  pivot_wider(
    names_from = method,
    values_from = sum
  ) |> 
  left_join(
    na_df_long |> 
      summarise(
        n_observed = sum(!is.na(gdp_b)),
        .by = country_name
      ),
    by = "country_name"
  )
```

so the table lets us see the three choices together: when we use na.rm = FALSE, a column that has values and NAs is always NA. But perhaps we are interested in preserving the NAs only in cases where all rows are NA. then, we need to modify the function. It is also often useful to show the amount of missingness itself. Otherwise, two columns can have the same sum while one of them is based on much less data.

# Plotting
There are much better guides than whatever I'll write, but some important things - 
a) axis titles are almost always unnecessary, unreadable, ugly and distracting (especially y axis titles). The same information can usually be conveyed using the main plot title

```{r plot-gdp-pc-raw}
eu27_countries <- c(
  "AT", "BE", "BG", "HR", "CY", "CZ", "DK", "EE", "FI", "FR",
  "DE", "GR", "HU", "IE", "IT", "LV", "LT", "LU", "MT", "NL",
  "PL", "PT", "RO", "SK", "SI", "ES", "SE"
)
highlight_countries <- c("IE", "LU", "SE")

plot_df <- country_panel |> 
  filter(
    iso2 %in% eu27_countries,
    year >= 2000,
    year <= 2024
  ) |> 
  mutate(
    gdp_pc = gdp_b * 1000 / pop_m,
    highlighted = if_else(iso2 %in% highlight_countries, iso2, "Other countries")
  )

plot_df |> 
  ggplot() + 
  geom_line(aes(x = year, y = gdp_pc, color = iso2)) +
  scale_x_continuous(breaks = seq(2000, 2024, 4), limits = c(2000, 2024)) +
  scale_y_continuous(limits = c(0, NA), labels = label_dollar())
```
Very minimal adjustments make it a lot cleaner:
```{r plot-gdp-pc-clean}
plot_df |> 
  ggplot() + 
  geom_line(aes(x = year, y = gdp_pc, color = iso2)) + 
  scale_x_continuous(breaks = seq(2000, 2024, 4), limits = c(2000, 2024)) +
  scale_y_continuous(limits = c(0, NA), labels = label_dollar()) +
  theme_minimal() + 
  theme(axis.title = element_blank()) + 
  labs(title = "GDP per capita across EU countries")
```

The legend can also often be avoided by placing the labels directly near the lines. The plot can also benefit from clearer colours (that are more colour-blind friendly). Say that we want to focus on Ireland's rise from Sweden's level of GDP per capita to nearly Luxembourg's, then these could be higlighted with the other EU countries having grey :

```{r plot-gdp-pc-highlighted}
plot_colors <- c(
  "IE" = "#0072B2",
  "LU" = "#D55E00",
  "SE" = "#009E73",
  "Other countries" = "#b8bec6"
)

line_labels <- plot_df |> 
  filter(iso2 %in% highlight_countries) |> 
  summarise(
    gdp_pc = gdp_pc[which.max(year)],
    year = max(year),
    label = country_name[which.max(year)],
    .by = iso2
  )

plot_df |> 
  ggplot(aes(x = year, y = gdp_pc, group = iso2)) + 
  geom_line(aes(color = highlighted), linewidth = 0.9, alpha = 0.85) + 
  geom_text(
    data = line_labels,
    aes(label = iso2, color = iso2),
    hjust = 0,
    nudge_x = 0.25,
    show.legend = FALSE
  ) +
  scale_color_manual(values = plot_colors) +
  scale_x_continuous(breaks = seq(2000, 2024, 4), limits = c(2000, 2025.5)) +
  scale_y_continuous(limits = c(0, NA), labels = label_dollar()) +
  theme_minimal() + 
  theme(
    axis.title = element_blank(),
    legend.position = "none",
    plot.margin = margin(5.5, 60, 5.5, 5.5)
  ) + 
  coord_cartesian(clip = "off") +
  labs(title = "Ireland spiked toward the EU's richest")
```

Finally, a more polished version could use background colours, cleaner axis, using the actual names as labels and so on:

```{r plot-gdp-pc-final}
plot_df |> 
  ggplot(aes(x = year, y = gdp_pc, group = iso2)) + 
  geom_line(aes(color = highlighted), linewidth = 0.9, alpha = 0.8) + 
  geom_text(
    data = line_labels,
    aes(label = label, color = iso2),
    hjust = 0,
    nudge_x = 0.25,
    fontface = "bold",
    show.legend = FALSE
  ) +
  scale_color_manual(values = plot_colors) +
  scale_x_continuous(breaks = seq(2000, 2024, 4), limits = c(2000, 2025.5)) +
  scale_y_continuous(limits = c(0, NA), labels = label_dollar()) +
  labs(
    title = "Ireland spiked toward the EU's richest",
    subtitle = "GDP per capita, constant 2015 US dollars",
    caption = "Source: The World Bank"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.title = element_blank(),
    legend.position = "none",
    plot.background = element_rect(fill = "#f7f8f3", color = NA),
    panel.background = element_rect(fill = "#f7f8f3", color = NA),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    plot.caption = element_text(color = "#6b7280"),
    plot.margin = margin(5.5, 70, 5.5, 5.5)
  ) +
  coord_cartesian(clip = "off")
```


# Spatial Analysis

A note (that I quite often ignore) - gpkg is recommended over shapefile [why]

### What to use

For me, there are mainly three types of tools that are useful in spatial analysis.

| Type | Examples | Main advantages | Main drawbacks |
|---|---|---|---|
| Programming languages | R with `sf` and `terra`; Python with `geopandas`, `rasterio`, `xarray`, etc. | Reproducible workflows; easy to combine spatial work with data cleaning, modeling, and plotting; good for local vector/raster operations | Local storage and memory can become a problem; worse for some visual validation |
| GIS software | QGIS, ArcGIS | Inspection, validation, manual exploration, and cartographic output | requires local storage; less convenient for some analysis workflows |
| Google Earth Engine | GEE Code Editor, GEE Python API | Hosted data catalogs; convenient raster and remote-sensing operations; interactive map checking; avoids many local storage problems | Less intuitive coding and debugging; quotas and server-side limits; less convenient for some reproducible project workflows |

**Programming languages** - can do most of the work well. Running things is rather fast for most cases and visualizations (e.g., `ggplot2::geom_sf()`) are decent. Mainly, they are very convenient when spatial work is a part of a larger data construction or analysis workflow. 

**GIS software** - ArcGIS is powerful but not free; QGIS is very good for visualizations and debugging or validating data (e.g., examine whether a spatial join worked properly or matched the wrong places). It is also possible to program there, but I find the programming less convenient than `sf` in R (decide for yourself, though; perhaps starting out with QGIS is easier), and it creates an additional layer of workflow (save the data, load it elsewhere, etc.). I personally like working with R though, especially with AI agents helping write and debug code. 

**GEE** - useful mainly, in my opinion, when the data already exists in the Earth Engine data catalog (https://developers.google.com/earth-engine/datasets) or the GEE community catalog (https://gee-community-catalog.org/). For example, using Hansen forest loss data or built-up data from GHSL is easier when one can avoid downloading and storing the data, and instead we'd run the relevant operations there. It can also be advantageous for raster or remote-sensing operations, such as filtering large image collections by date/cloud cover/location, combining multiple images, and using reducers over space or time. However, this is less of a reason to use GEE for every heavy spatial process now that there are quotas for noncommercial users. It is also good for validation and checking layers on a map. The main drawback of GEE, beyond quotas, is the coding experience: the Code Editor uses the Earth Engine JavaScript API, many objects are server-side, and errors can be opaque. GEE now has a Gemini assistant in the Code Editor (which I haven't tried yet), which requires a Gemini API key, so it is not necessarily unlimited/free. In terms of workflows, it can be annoying, although it's possible to use the Python API to keep all the code in a single project with other R/Python/Stata codes.

When unsure which tool to use, AIs would probably be useful to help you judge, and now the costs of writing a sample code in R and examining its efficiency are so low that it's probably worth starting from it.

### spatial matching

discuss here how to intersect between spatial data and some of the relevant considerations (whether technical - st_intersection vs st_nearest_feature, but also raster vs vector, what do to in cases of multiple intersections etc)

### polygons vs points
sometimes we may want to work with single points in a given unit, rather than the full polygon. that's especially the case if 1) the analysis is very heavy (e.g., calculating road path from unit x to y); 2) we are interested in the role of specific areas within the unit (e.g., the main city), 3) fill
if we chooes to use points, how do we select the point? it depends obviously on what we want to achieve.
centroid is a popular solution, but it suffers from various drawbacks. 
- it is not necessarily inside the unit. think of units with weird shapes (vietnam), that are composed of many distinct geometries (multiple-island countries, like the philippines) or with holes inside (south africa). if we're calculating the distance of centroid of country x to that of y, and country y's centroid is inside that of x's, it's not ideal...
- one solution for that would be to use `st_point_on_surface` from the sf package. it guarantees that the point would be on the surface of the unit, and generally is around its middle. however, for certain cases, it might still not capture something we're interested in. I was working with an admin unit that included a city in the mainland, and an island very far from the mainland. The island was larger, so `st_point_on_surface` has led to a point in the island. In some cases perhaps it's fine, but if we're interested in anything that involves population - then it really does not capture what we want

```{r plot-centroid-surface-points, echo=FALSE}
normal_polygon <- st_polygon(list(
  matrix(
    c(
      0.2, 0.1,
      4.1, 0.0,
      4.4, 2.2,
      3.6, 4.0,
      1.3, 4.3,
      0.0, 2.5,
      0.2, 0.1
    ),
    ncol = 2,
    byrow = TRUE
  )
))

hole_polygon <- st_polygon(list(
  matrix(c(0, 0, 4, 0, 4, 4, 0, 4, 0, 0), ncol = 2, byrow = TRUE),
  matrix(c(1.2, 1.2, 1.2, 2.8, 2.8, 2.8, 2.8, 1.2, 1.2, 1.2), ncol = 2, byrow = TRUE)
))

schematic_units <- st_sf(
  case = c("compact polygon", "polygon with hole"),
  geometry = st_sfc(normal_polygon, hole_polygon)
)

centroids <- schematic_units |>
  st_centroid() |>
  mutate(point_type = "centroid")

surface_points <- schematic_units |>
  st_point_on_surface() |>
  mutate(point_type = "point on surface")

bind_rows(centroids, surface_points) |>
  ggplot() +
  geom_sf(data = schematic_units, fill = "#e5e7eb", color = "#4b5563") +
  geom_sf(aes(color = point_type), size = 2.8) +
  facet_wrap(~case) +
  scale_color_manual(values = c("centroid" = "#D55E00", "point on surface" = "#0072B2")) +
  coord_sf(datum = NA) +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.title = element_blank()
  )
```

- what about population-weighted centroid? we can use data like GHS-POP to do it (recommended to do it in GEE). However, 1) it might still be outside of the unit, 2) say that there are several population centres in different areas. the pop-weighted centroid is probably going to be between them, which might be good but it'll likely reflect a (relatively) remote point. which might be fine but not always!

```{r plot-population-centroid-points, echo=FALSE}
population_unit <- st_sf(
  geometry = st_sfc(normal_polygon)
)

mock_population_grid <- st_sf(
  geometry = st_make_grid(population_unit, n = c(10, 10))
) |>
  st_intersection(population_unit)

grid_centers <- st_coordinates(st_centroid(mock_population_grid))

mock_population_grid <- mock_population_grid |>
  mutate(
    x_coord = grid_centers[, 1],
    y_coord = grid_centers[, 2],
    population = round(
      15 +
        950 * exp(-((x_coord - 1.0)^2 + (y_coord - 3.0)^2) / 0.18) +
        240 * exp(-((x_coord - 3.2)^2 + (y_coord - 1.0)^2) / 0.14) +
        180 * exp(-((x_coord - 3.4)^2 + (y_coord - 3.1)^2) / 0.16) +
        130 * exp(-((x_coord - 2.0)^2 + (y_coord - 1.9)^2) / 0.12)
    )
  )

population_weighted_point <- mock_population_grid |>
  st_drop_geometry() |>
  summarise(
    x_coord = weighted.mean(x_coord, population),
    y_coord = weighted.mean(y_coord, population)
  ) |>
  st_as_sf(coords = c("x_coord", "y_coord")) |>
  mutate(point_type = "population-weighted centroid")

highest_population_point <- mock_population_grid |>
  slice_max(population, n = 1, with_ties = FALSE) |>
  st_centroid() |>
  mutate(point_type = "highest population cell")

population_centroid <- population_unit |>
  st_centroid() |>
  mutate(point_type = "centroid")

population_surface_point <- population_unit |>
  st_point_on_surface() |>
  mutate(point_type = "point on surface")

population_summary_points <- bind_rows(
  population_centroid,
  population_surface_point,
  population_weighted_point,
  highest_population_point
)

population_summary_labels <- population_summary_points |>
  mutate(
    point_x = st_coordinates(geometry)[, 1],
    point_y = st_coordinates(geometry)[, 2],
    label_x = case_when(
      point_type == "highest population cell" ~ point_x - 0.16,
      point_type == "population-weighted centroid" ~ point_x + 0.18,
      point_type == "centroid" ~ point_x + 0.18,
      TRUE ~ point_x - 0.16
    ),
    label_y = case_when(
      point_type == "highest population cell" ~ point_y + 0.20,
      point_type == "population-weighted centroid" ~ point_y - 0.18,
      point_type == "centroid" ~ point_y + 0.18,
      TRUE ~ point_y - 0.20
    )
  ) |>
  st_drop_geometry()

ggplot() +
  geom_sf(
    data = mock_population_grid,
    aes(fill = population),
    color = "white",
    linewidth = 0.2
  ) +
  geom_sf(data = population_unit, fill = NA, color = "#4b5563", linewidth = 0.5) +
  geom_sf(
    data = population_summary_points,
    aes(color = point_type),
    size = 3.4
  ) +
  geom_label(
    data = population_summary_labels,
    aes(x = label_x, y = label_y, label = point_type),
    size = 2.8,
    linewidth = 0.15,
    label.padding = unit(0.08, "lines")
  ) +
  scale_fill_gradient(low = "white", high = "#b91c1c") +
  scale_color_manual(
    values = c(
      "centroid" = "#D55E00",
      "point on surface" = "#0072B2",
      "population-weighted centroid" = "#009E73",
      "highest population cell" = "black"
    )
  ) +
  coord_sf(datum = NA, expand = TRUE) +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  )
```

- similarly to the last suggestion, can use 'the most populated point'. for certain things it might be preferable (for others, not as much). one thing to consider here, as many spatial analysis, is the resolution. the most populated 100m place in a country could theoretically be a small town that has an extremely dense point. so aggregating by 1/5/10km might be preferable, but examine the outputs and see whether this captures what you're interested in

### finding a 'representative' category
there are various spatial datasets that might involve finding some balance between various categories. mines datasets, for example, sometimes have each individual mine as a row, or would have a value of 'suitability for a given commodity' per cell. depends on what we want to have, we might want to aggregate the data per unit. should we sum the number of distinct mines? use the commodity with the highest value of suitability? no obvious answer here but worth thinking about those...

# Analysis
log() of low numbers is pretty useless. not unknown, but worth paying attention to

## shift share

# Debugging and validating results

- row counts before and after each important step
- duplicate keys
- unmatched observations
- impossible values
- missingness summaries
- graphs
- edge cases
- compare aggregates to an external source
- inspect examples rather than only summary statistics

# constructing workflows

- raw, intermediate, and output data
- script order
- naming conventions
- expensive steps and cached outputs
- logs
- seeds
- final checklist before sending results

# AI use

# Specific datasets

## WDPA (World Database on Protected Areas)

**Points** - The database includes protected areas (PAs) represented both as polygons and as points. For polygons, the boundaries of the PAs are available directly. Points are more ambiguous, as they can represent point-like features, or refer to larger protected areas for which the shape is unknown. The `REP_AREA` field can be used to approximate the size of these point records, but the point location is not necessarily the centroid of the area, and the shape of the area is unknown. Therefore, buffering points by their reported area may be useful as an approximation, but it is unlikely to recover the true shape of the PA.

**Overlaps** - When intersecting WDPA with other spatial data, for example to estimate the share of an area covered by PAs, need to account for overlapping of PAs. A given location can be covered by multiple PAs, so simply summing areas across records can overstate coverage. Thus, protected-area geometries may need to be unioned before calculating total covered area (perhaps by category, depending on the goal).

**Status year** - The time variable in WDPA is `STATUS_YR`, which records the last year of status change. This usually refers to the legal designation as a protected area. In checks against South African protected-area data from the late 1990s onward, most protected areas did not appear to change status after their initial designation during the observed period. This suggests that `STATUS_YR` can often be informative about designation timing, but it should still be used cautiously, especially for cases involving expansions, reclassification, proposed or established areas, and loss of status.

## IPUMS

# Suggestions for stuff to add / remove

- add a short opening paragraph about what kinds of RA work this is based on
- add one concrete example of a result validation workflow
- add a small checklist at the end
- maybe remove or shorten any examples that are too project-specific once the structure is clearer
