---
title: "RA lessons"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: hide
    css: styles.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# Introduction

this wouldnt speak much about measurement issues in datasets

```{r}
pacman::p_load(
  dplyr, labelled, sjlabelled, tidyr, ggplot2, scales, sf,
  here, readr
  )

options(scipen = 999)
```

# R

## style and speed

- I mainly used dplyr, so while data.table is faster, i find the syntax of it a bit less intuitive and prefer to work with the more piped structure of dplyr
- for faster processes while still using dplyr syntax, dtplyr is very useful. duckdb is way faster and presents another option that is useful for some cases
- highly recommend to examine updates to the dplyr package (and in general, to packages that are used often). 
- the new pipe (|>) is nicer and cleaner than the old magrittr `%>%` pipe

see the difference between coding styles:

```{r}
style_df <- tibble::tibble(
  iso2 = c("US", "US", "CN", "CN"),
  gdp_t = c(25, 26, 17, 18)
)

style_df
```

```{r}
style_df %>% 
  group_by(iso2) %>%
  summarise(
    avg_gdp = mean(gdp_t)
    ) %>%
  ungroup()
```

```{r}
style_df |> 
  summarise(
    avg_gdp = mean(gdp_t),
    .by = iso2
    ) 
```
Decide for yourself but I much prefer the latter.

## functions with the same names 

- there are some functions with identical names across packages. this is a source of some bugs and odd behaviour that is generally hard to understand or detect. 'lag' is a nice example of that. its base version from stats::lag is quite bad as it doesn't actually lag the variables; dplyr::lag (or data.table::shift(type = "lag") are much more reliable. raster::select might cause the code to break if dplyr::select is not mentioned explicitly (but here at least there wouldn't be a mysterious bug, unlike the 'lag' case). another example is 'copy\_labels', that exists in both 'sjlabelled' and 'labelled'. I once didn't examine the exact meaning of the packages, and incorrectly assumed that they would just copy the labels per variables while not changing the data.
But note that they do change the values and fill in values for the data with NA, but in an opposite direction depending on the order of the data, as the functions work in this way
```r
    sjlabelled::copy_labels(new, old)
    labelled::copy_labels(old, new)
    # to get them to be 'equivalent', one can e.g., use
    labelled::copy_labels_from()
    or set the exact parameters like labelled::copy_labels(to=new, from=old) 
```
example:

`old_labels` is a labelled version of the data, and `new_data` is a newer version of the data that we want to copy the labels into.

```{r include=FALSE}
old_labels <- tibble::tibble(
  x = labelled(c(1, 2), labels = c(no = 1, yes = 2)),
  y = labelled(c(10, 20), labels = c(low = 10, high = 20))
)

new_data <- tibble::tibble(
  x = c(NA_real_, 99),
  y = c(30, 40)
)
```

```{r}
old_labels
new_data
```

```{r}
labelled::copy_labels(old_labels, new_data)
sjlabelled::copy_labels(old_labels, new_data)
```

# Country level analysis

A lot of the analysis I have done was at the country level, or used countries as parent units. Thus, most of this text is going to use countries as example units, but here are some tips on dealing with country level data:

## Country names
Country names can be inconsistent between datasets. 'Ivory Coast' and 'Cote d'Ivoire' refer to the same country. While keeping the name can be useful for plots, tables and understanding the data, for joins it is much preferable to use country codes. ISO2 or ISO3 are great for that. Note, however, that even country codes may have some inconsistency, some examples -
1) South Sudan's ISO3 is usually 'SSD', but I've seen some datasets where it was 'SDS' 
2) Namibia's ISO2 is 'NA'. Depending on the function that reads the data, it may be read as NA

Normally the WDI includes data on aggregates (if you download directly with 'WDI::WDI(), use extra = TRUE')). For illustration, however, I have downloaded the data from the WDI website as csv. https://data.worldbank.org/indicator/SP.POP.TOTL and https://data.worldbank.org/indicator/NY.GDP.MKTP.KD

```{r}
selected_iso2 <- c("US", "CN", "EU")

pop <- read_csv(here("data", "API_SP.POP.TOTL_DS2_en_csv_v2_33112.csv"), skip = 4) |>
  pivot_longer(
    cols = matches("^\\d{4}$"),
    names_to = "year",
    values_to = "pop"
  ) |>
  transmute(
    iso3 = `Country Code`,
    country_name = `Country Name`,
    year = as.integer(year),
    pop
  )

gdp <- read_csv(here("data", "API_NY.GDP.MKTP.KD_DS2_en_csv_v2_34247.csv"), skip = 4) |>
  pivot_longer(
    cols = matches("^\\d{4}$"),
    names_to = "year",
    values_to = "gdp"
  ) |>
  transmute(
    iso3 = `Country Code`,
    country_name = `Country Name`,
    year = as.integer(year),
    gdp
  )

country_panel_all <- full_join(pop, gdp, by = c("iso3", "country_name", "year")) |>
  transmute(
    iso2 = countrycode::countrycode(iso3, "iso3c", "iso2c", warn = FALSE),
    iso3,
    country_name,
    year,
    gdp,
    gdp_b = round(gdp / 1e9, 5),
    gdp_t = round(gdp / 1e12, 5),
    pop
  ) |>
  filter(year >= 2000) |>
  arrange(iso2, year)

saveRDS(country_panel_all, here("data", "wdi_gdp_country_panel_all.rds"))
```

Now, this version does not already contain a column referring to whether a unit is an aggregate or not (the data actually exists in the metadata, but I'm not using it on purpose). There are many ways to deal with it, one (probably not the best one, but it highlights a potential problem) is using the 'countrycode' package. I converted all the ISO3s into ISO2, which leaves NAs at all the aggregate units in the newly created ISO2 column. However, if we're paying close attention, see that Kosovo also does not have ISO2 here.

```{r}
aggregate_or_unmatched_names <- country_panel_all |> 
  filter(is.na(iso2)) |> 
  distinct(country_name) |> 
  pull(country_name)

cat(paste(aggregate_or_unmatched_names, collapse = ", "))
```

Thus, I'd manually add the ISO3 for Kosovo (and the EU, as I want to use it).

```{r}
country_panel <- country_panel_all |> 
  mutate(
    iso2 = case_when(
      iso3 == "XKX" ~ "XK",
      iso3 == "EUU" ~ "EU", 
      TRUE ~ iso2
      )
  ) |> 
  filter(!is.na(iso2))
```

## country aggregates

In some datasets there may be aggregates (e.g., "Middle East"). In some datasets there are variables that refer to those, but not always, and the names of the 'aggregate' countries might be different across different datasets (e.g., in two different datasets of FAOSTAT, there is in one dataset (https://www.fao.org/faostat/en/#data/FBSH) an aggregated region of 'Low Income Food Deficit Countries', but in another (https://www.fao.org/faostat/en/#data/QCL) this appears as 'Low Income Food Deficit Countries (LIFDCs)'. Given that there is not always an ISO2/3 equivalent for each aggregate, one needs to pay attention to those cases. 

<!-- What if we want to aggregate multiple rows. I've removed before the EU as an aggregate, but let's assume that we'd want to construct it from scratch: -->
<!-- ```{r} -->
<!-- eu_df <- country_panel |>  -->
<!--   mutate( -->
<!--     iso2 = if_else(iso2 %in% eu_iso2, "EU", iso2), -->
<!--     country_name = if_else(iso2 == "EU", "European Union", country_name) -->
<!--   ) |> -->
<!--   summarise( -->
<!--     gdp = if (all(is.na(gdp))) NA_real_ else sum(gdp, na.rm = TRUE), -->
<!--     gdp_b = round(gdp / 1e9, 1), -->
<!--     gdp_t = round(gdp / 1e12, 1), -->
<!--     pop = sum(pop, na.rm = TRUE), -->
<!--     .by = c(iso2, country_name, year) -->
<!--   ) -->
<!-- ``` -->

<!-- # eu_comparison <- eu_df |>  -->
<!-- #   filter(iso2 == "EU") |>  -->
<!-- #   select(year, eu_constructed_gdp_t = gdp_t, eu_constructed_pop = pop) |> -->
<!-- #   left_join( -->
<!-- #     country_panel_all |>  -->
<!-- #       filter(iso3 == "EUU") |>  -->
<!-- #       select(year, eu_wdi_gdp_t = gdp_t, eu_wdi_pop = pop), -->
<!-- #     by = "year" -->
<!-- #   ) |> -->
<!-- #   mutate( -->
<!-- #     gdp_diff_t = round(eu_constructed_gdp_t - eu_wdi_gdp_t, 2), -->
<!-- #     pop_diff = eu_constructed_pop - eu_wdi_pop -->
<!-- #   ) -->
<!-- #  -->
<!-- # eu_comparison -->

<!-- Importantly for cases like this, ideally we'd want to verify that the values make sense compared to some reference data. For instance, I would try to find data on the aggregate (EU, in this case) from another source - IMF, Eurostat, etc, asking an AI whether it looks plausible etc.  -->


## Geographies consistency over time

Various geographical units of observations do not have consistent boundaries over time. Countries change their borders, gain independence or split; admin units sometimes change have changed their borders. Different datasets treat those changes differently - perhaps a value in a given year refers to the current border or the old one. 
When working with country-level data, for example, pay attention to countries whose borders have changed in the period of analysis (or recently!). 'Sudan' might reflect the current state of Sudan, or also the values of South Sudan. Montenegro might be part of Yugoslavia, Serbia and Montenegro, or Montenegro; the same is true for Kosovo (despite not being mentioned in the name of 'Serbia and Montenegro'), although Kosovo might not even exist in certain datasets. Taiwan might be referred to in different ways, depending on the dataset.

This, for example, looks rather fine - there are no obvious spikes in the data around the year of independence:
```{r}
country_panel |>
  filter(iso2 %in% c("SD", "SS"), year %in% 2006:2014) |>
  ggplot() + 
  geom_line(aes(x = year, y = pop, group = country_name, color = country_name))

```

# NAs
Assume that we have a dataset with some missing values:
```{r}
na_df <- country_panel |>
  filter(
    iso2 %in% c("CU", "ER", "BO"), 
    year %in% 2023:2025
    ) |>
  select(year, country_name, gdp_b) |> 
  pivot_wider(
    names_from = country_name,
    values_from = c(gdp_b)
  )

na_df
```
We want to aggregate that data. The main options are 1) na.rm = TRUE, which removes the NAs before calculating, such that sum(df$x, na.rm = TRUE) would still return a value. For sums this can look like treating NA as 0, and if all values are NA it returns 0, but for means it also changes the denominator; 2) na.rm = FALSE, which forces the aggregation to have NA if any of the values is NA 

Especially if we use a loop or across to aggregate over several variables or units, it can be a problem
Say that we want to sum gdp for each of those three countries in this period.  
```{r}
na_df_long <- na_df |> 
  pivot_longer(
    -year,
    names_to = "country_name",
    values_to = "gdp_b"
  )

na_rm_options <- c("na.rm = TRUE", "na.rm = FALSE", "NA only if all missing")

bind_rows(
  na_df_long |> 
    summarise(
      method = "na.rm = TRUE",
      sum = sum(gdp_b, na.rm = TRUE),
      .by = country_name
    ),
  na_df_long |> 
    summarise(
      method = "na.rm = FALSE",
      sum = sum(gdp_b, na.rm = FALSE),
      .by = country_name
    ),
  na_df_long |> 
    summarise(
      method = "NA only if all missing",
      sum = if (all(is.na(gdp_b))) NA_real_ else sum(gdp_b, na.rm = TRUE),
      .by = country_name
    )
  ) |> 
  arrange(country_name, match(method, na_rm_options)) |> 
  pivot_wider(
    names_from = method,
    values_from = sum
  ) |> 
  left_join(
    na_df_long |> 
      summarise(
        n_observed = sum(!is.na(gdp_b)),
        .by = country_name
      ),
    by = "country_name"
  )
```

so the table lets us see the three choices together: when we use na.rm = FALSE, a column that has values and NAs is always NA. But perhaps we are interested in preserving the NAs only in cases where all rows are NA. then, we need to modify the function. It is also often useful to show the amount of missingness itself. Otherwise, two columns can have the same sum while one of them is based on much less data.

# Plotting
There are much better guides than whatever I'll write, but some important things - 
a) axis titles are almost always unnecessary, unreadable, ugly and distracting (especially y axis titles). The same information can usually be conveyed using the main plot title

```{r}
eu27_countries <- c(
  "AT", "BE", "BG", "HR", "CY", "CZ", "DK", "EE", "FI", "FR",
  "DE", "GR", "HU", "IE", "IT", "LV", "LT", "LU", "MT", "NL",
  "PL", "PT", "RO", "SK", "SI", "ES", "SE"
)
highlight_countries <- c("IE", "LU", "SE")

plot_df <- country_panel |> 
  filter(
    iso2 %in% eu27_countries,
    year >= 2000,
    year <= 2024
  ) |> 
  mutate(
    gdp_pc = gdp / pop,
    highlighted = if_else(iso2 %in% highlight_countries, iso2, "Other countries")
  )

plot_df |> 
  ggplot() + 
  geom_line(aes(x = year, y = gdp_pc, color = iso2)) +
  scale_x_continuous(breaks = seq(2000, 2024, 4), limits = c(2000, 2024)) +
  scale_y_continuous(limits = c(0, NA), labels = label_dollar())
```
Very minimal adjustments make it a lot cleaner:
```{r}
plot_df |> 
  ggplot() + 
  geom_line(aes(x = year, y = gdp_pc, color = iso2)) + 
  scale_x_continuous(breaks = seq(2000, 2024, 4), limits = c(2000, 2024)) +
  scale_y_continuous(limits = c(0, NA), labels = label_dollar()) +
  theme_minimal() + 
  theme(axis.title = element_blank()) + 
  labs(title = "GDP per capita across EU countries")
```

the legend can also often be avoided by placing the labels directly near the lines. The plot can also benefit from clearer colours (that are more colour-blind friendly). Say that we want to focus on Ireland, Luxembourg, and Sweden
```{r}
plot_colors <- c(
  "IE" = "#0072B2",
  "LU" = "#D55E00",
  "SE" = "#009E73",
  "Other countries" = "#b8bec6"
)

line_labels <- plot_df |> 
  filter(iso2 %in% highlight_countries) |> 
  summarise(
    gdp_pc = gdp_pc[which.max(year)],
    year = max(year),
    label = country_name[which.max(year)],
    .by = iso2
  )

plot_df |> 
  ggplot(aes(x = year, y = gdp_pc, group = iso2)) + 
  geom_line(aes(color = highlighted), linewidth = 0.9, alpha = 0.85) + 
  geom_text(
    data = line_labels,
    aes(label = iso2, color = iso2),
    hjust = 0,
    nudge_x = 0.25,
    show.legend = FALSE
  ) +
  scale_color_manual(values = plot_colors) +
  scale_x_continuous(breaks = seq(2000, 2024, 4), limits = c(2000, 2025.5)) +
  scale_y_continuous(limits = c(0, NA), labels = label_dollar()) +
  theme_minimal() + 
  theme(
    axis.title = element_blank(),
    legend.position = "none",
    plot.margin = margin(5.5, 60, 5.5, 5.5)
  ) + 
  coord_cartesian(clip = "off") +
  labs(title = "Ireland spiked toward the EU's richest")
```

and a more final version could look something like this:
```{r}
plot_df |> 
  ggplot(aes(x = year, y = gdp_pc, group = iso2)) + 
  geom_line(aes(color = highlighted), linewidth = 0.9, alpha = 0.8) + 
  geom_text(
    data = line_labels,
    aes(label = label, color = iso2),
    hjust = 0,
    nudge_x = 0.25,
    fontface = "bold",
    show.legend = FALSE
  ) +
  scale_color_manual(values = plot_colors) +
  scale_x_continuous(breaks = seq(2000, 2024, 4), limits = c(2000, 2025.5)) +
  scale_y_continuous(limits = c(0, NA), labels = label_dollar()) +
  labs(
    title = "Ireland spiked toward the EU's richest",
    subtitle = "GDP per capita, constant 2015 US dollars",
    caption = "Source: The World Bank"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.title = element_blank(),
    legend.position = "none",
    plot.background = element_rect(fill = "#f7f8f3", color = NA),
    panel.background = element_rect(fill = "#f7f8f3", color = NA),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    plot.caption = element_text(color = "#6b7280"),
    plot.margin = margin(5.5, 70, 5.5, 5.5)
  ) +
  coord_cartesian(clip = "off")
```


# Spatial Analysis

- sf is great
- gpkg is recommended over shapefile 
- gee is great, but for some stuff r does the job done - gee is in my opinion mainly advantageous if the data is already there, if certain operations (geometric operations for example) are required, or heavier processes (although with the new limits, less so). 

### polygons vs points
sometimes we may want to work with single points in a given unit, rather than the full polygon. that's especially the case if 1) the analysis is very heavy (e.g., calculating road path from unit x to y); 2) we are interested in the role of specific areas within the unit (e.g., the main city), 3) fill
if we chooes to use points, how do we select the point? it depends obviously on what we want to achieve.
centroid is a popular solution, but it suffers from various drawbacks. 
- it is not necessarily inside the unit. think of units with weird shapes (vietnam), that are composed of many distinct geometries (multiple-island countries, like the philippines) or with holes inside (south africa). if we're calculating the distance of centroid of country x to that of y, and country y's centroid is inside that of x's, it's not ideal...
- one solution for that would be to use `st_point_on_surface` from the sf package. it guarantees that the point would be on the surface of the unit, and generally is around its middle. however, for certain cases, it might still not capture something we're interested in. I was working with an admin unit that included a city in the mainland, and an island very far from the mainland. The island was larger, so `st_point_on_surface` has led to a point in the island. In some cases perhaps it's fine, but if we're interested in anything that involves population - then it really does not capture what we want

```{r echo=FALSE}
normal_polygon <- st_polygon(list(
  matrix(
    c(
      0.2, 0.1,
      4.1, 0.0,
      4.4, 2.2,
      3.6, 4.0,
      1.3, 4.3,
      0.0, 2.5,
      0.2, 0.1
    ),
    ncol = 2,
    byrow = TRUE
  )
))

hole_polygon <- st_polygon(list(
  matrix(c(0, 0, 4, 0, 4, 4, 0, 4, 0, 0), ncol = 2, byrow = TRUE),
  matrix(c(1.2, 1.2, 1.2, 2.8, 2.8, 2.8, 2.8, 1.2, 1.2, 1.2), ncol = 2, byrow = TRUE)
))

schematic_units <- st_sf(
  case = c("compact polygon", "polygon with hole"),
  geometry = st_sfc(normal_polygon, hole_polygon)
)

centroids <- schematic_units |>
  st_centroid() |>
  mutate(point_type = "centroid")

surface_points <- schematic_units |>
  st_point_on_surface() |>
  mutate(point_type = "point on surface")

bind_rows(centroids, surface_points) |>
  ggplot() +
  geom_sf(data = schematic_units, fill = "#e5e7eb", color = "#4b5563") +
  geom_sf(aes(color = point_type), size = 2.8) +
  facet_wrap(~case) +
  scale_color_manual(values = c("centroid" = "#D55E00", "point on surface" = "#0072B2")) +
  coord_sf(datum = NA) +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.title = element_blank()
  )
```

- what about population-weighted centroid? we can use data like GHS-POP to do it (recommended to do it in GEE). However, 1) it might still be outside of the unit, 2) say that there are several population centres in different areas. the pop-weighted centroid is probably going to be between them, which might be good but it'll likely reflect a (relatively) remote point. which might be fine but not always!

```{r echo=FALSE}
population_unit <- st_sf(
  geometry = st_sfc(normal_polygon)
)

mock_population_grid <- st_sf(
  geometry = st_make_grid(population_unit, n = c(10, 10))
) |>
  st_intersection(population_unit)

grid_centers <- st_coordinates(st_centroid(mock_population_grid))

mock_population_grid <- mock_population_grid |>
  mutate(
    x_coord = grid_centers[, 1],
    y_coord = grid_centers[, 2],
    population = round(
      15 +
        950 * exp(-((x_coord - 1.0)^2 + (y_coord - 3.0)^2) / 0.18) +
        240 * exp(-((x_coord - 3.2)^2 + (y_coord - 1.0)^2) / 0.14) +
        180 * exp(-((x_coord - 3.4)^2 + (y_coord - 3.1)^2) / 0.16) +
        130 * exp(-((x_coord - 2.0)^2 + (y_coord - 1.9)^2) / 0.12)
    )
  )

population_weighted_point <- mock_population_grid |>
  st_drop_geometry() |>
  summarise(
    x_coord = weighted.mean(x_coord, population),
    y_coord = weighted.mean(y_coord, population)
  ) |>
  st_as_sf(coords = c("x_coord", "y_coord")) |>
  mutate(point_type = "population-weighted centroid")

highest_population_point <- mock_population_grid |>
  slice_max(population, n = 1, with_ties = FALSE) |>
  st_centroid() |>
  mutate(point_type = "highest population cell")

population_centroid <- population_unit |>
  st_centroid() |>
  mutate(point_type = "centroid")

population_surface_point <- population_unit |>
  st_point_on_surface() |>
  mutate(point_type = "point on surface")

population_summary_points <- bind_rows(
  population_centroid,
  population_surface_point,
  population_weighted_point,
  highest_population_point
)

population_summary_labels <- population_summary_points |>
  mutate(
    point_x = st_coordinates(geometry)[, 1],
    point_y = st_coordinates(geometry)[, 2],
    label_x = case_when(
      point_type == "highest population cell" ~ point_x - 0.16,
      point_type == "population-weighted centroid" ~ point_x + 0.18,
      point_type == "centroid" ~ point_x + 0.18,
      TRUE ~ point_x - 0.16
    ),
    label_y = case_when(
      point_type == "highest population cell" ~ point_y + 0.20,
      point_type == "population-weighted centroid" ~ point_y - 0.18,
      point_type == "centroid" ~ point_y + 0.18,
      TRUE ~ point_y - 0.20
    )
  ) |>
  st_drop_geometry()

ggplot() +
  geom_sf(
    data = mock_population_grid,
    aes(fill = population),
    color = "white",
    linewidth = 0.2
  ) +
  geom_sf(data = population_unit, fill = NA, color = "#4b5563", linewidth = 0.5) +
  geom_sf(
    data = population_summary_points,
    aes(color = point_type),
    size = 3.4
  ) +
  geom_label(
    data = population_summary_labels,
    aes(x = label_x, y = label_y, label = point_type),
    size = 2.8,
    linewidth = 0.15,
    label.padding = unit(0.08, "lines")
  ) +
  scale_fill_gradient(low = "white", high = "#b91c1c") +
  scale_color_manual(
    values = c(
      "centroid" = "#D55E00",
      "point on surface" = "#0072B2",
      "population-weighted centroid" = "#009E73",
      "highest population cell" = "#CC79A7"
    )
  ) +
  coord_sf(datum = NA, expand = TRUE) +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  )
```

- similarly to the last suggestion, can use 'the most populated point'. for certain things it might be preferable (for others, not as much). one thing to consider here, as many spatial analysis, is the resolution. the most populated 100m place in a country could theoretically be a small town that has an extremely dense point. so aggregating by 1/5/10km might be preferable, but examine the outputs and see whether this captures what you're interested in

### finding a 'representative' category
there are various spatial datasets that might involve finding some balance between various categories. mines datasets, for example, sometimes have each individual mine as a row, or would have a value of 'suitability for a given commodity' per cell. depends on what we want to have, we might want to aggregate the data per unit. should we sum the number of distinct mines? use the commodity with the highest value of suitability? no obvious answer here but worth thinking about those...

# empirical
log() of low numbers is pretty useless. not unknown, but worth paying attention to

## shift share

# Debugging and validating results

# Specific datasets

## WDPA (protected areas)

## IPUMS

# Suggestions for stuff to add / remove

- add a short opening paragraph about what kinds of RA work this is based on
- add one concrete example of a result validation workflow
- add a small checklist at the end
- maybe remove or shorten any examples that are too project-specific once the structure is clearer
