---
title: "RA lessons"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: hide
    css: styles.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# Introduction

this wouldnt speak much about measurement issues in datasets

```{r load-packages}
pacman::p_load(
  dplyr, labelled, sjlabelled, tidyr, ggplot2, scales, sf,
  here, readr, dtplyr, duckdb, DBI, rnaturalearth
  )

options(scipen = 999)
```

maybe say something about whether the code should be aimed at humans or ai - recommend something that is at least readable and debuggable by humans, even if not the cleanest. maybe use some example of a way that ai would write a code 'efficiently' vs humans - perhaps loop is a good example? or even just the number of messages etc

# Working in R

## style and speed

I mainly used `dplyr`, due to a combination of preference for its syntax and the piped structure, and being used to it. 


```{r make-style-example}
example_df <- tibble::tibble(
  iso2 = c("US", "US", "CN", "CN"),
  gdp_t = c(25, 26, 17, 18)
)
```

- highly recommend to examine updates to the dplyr package (and in general, to packages that are used often). 
- the new pipe (|>) is nicer and cleaner than the old magrittr `%>%` pipe

see the difference between coding styles:

```{r old-pipe-style, results='hide', class.source='fold-show'}
example_df %>% 
  group_by(iso2) %>%
  summarise(
    avg_gdp = mean(gdp_t)
    ) %>%
  ungroup()
```

```{r native-pipe-style, results='hide', class.source='fold-show'}
example_df |> 
  summarise(
    avg_gdp = mean(gdp_t),
    .by = iso2
    ) 
```
Decide for yourself but I much prefer the latter.

`data.table` is faster, but for most processes the marginal improvement is rather small. 

For heavier processes, one can use `dtplyr`. For smaller processes, it seems better to use `dplyr` regularly, as the setup costs may take longer than the actual calculation.

```{r benchmark-dplyr-dtplyr}
set.seed(23072026)

speed_df <- tibble::tibble(
  iso2 = sample(sprintf("C%03d", 1:200), 3000000, replace = TRUE),
  year = sample(2000:2024, 3000000, replace = TRUE),
  gdp_t = runif(3000000, 1, 100),
  pop_m = runif(3000000, 1, 1500)
)

dplyr_elapsed <- system.time({
  dplyr_result <- speed_df |>
    mutate(
      gdp_pc = gdp_t * 1000000 / pop_m,
      large = gdp_t > 50
    ) |>
    filter(large, year >= 2010) |>
    summarise(
      avg_gdp_pc = mean(gdp_pc),
      total_gdp = sum(gdp_t),
      n = n(),
      .by = c(iso2, year)
      )
})[["elapsed"]]

cat("dplyr elapsed:", round(dplyr_elapsed, 3), "seconds\n")

dtplyr_elapsed <- system.time({
  dtplyr_result <- speed_df |>
    lazy_dt() |>
    mutate(
      gdp_pc = gdp_t * 1000000 / pop_m,
      large = gdp_t > 50
    ) |>
    filter(large, year >= 2010) |>
    summarise(
      avg_gdp_pc = mean(gdp_pc),
      total_gdp = sum(gdp_t),
      n = n(),
      .by = c(iso2, year)
      ) |>
    as_tibble()
})[["elapsed"]]

cat("dtplyr elapsed:", round(dtplyr_elapsed, 3), "seconds\n")
```

`duckdb` is another fast option, especially when the data is too large to comfortably keep in memory. It requires SQL syntax, however.

```{r benchmark-duckdb}
con <- dbConnect(duckdb())
duckdb_register(con, "speed_df", speed_df)

duckdb_elapsed <- system.time({
  duckdb_result <- dbGetQuery(
    con,
    "
    SELECT
      iso2,
      year,
      AVG(gdp_t * 1000000 / pop_m) AS avg_gdp_pc,
      SUM(gdp_t) AS total_gdp,
      COUNT(*) AS n
    FROM speed_df
    WHERE gdp_t > 50
      AND year >= 2010
    GROUP BY iso2, year
    "
  )
})[["elapsed"]]

cat("duckdb elapsed:", round(duckdb_elapsed, 3), "seconds\n")

dbDisconnect(con, shutdown = TRUE)
```

## functions with the same names 

Some R functions from different packages have the identical names. This is the source of some bugs and unexpected behaviour, that is sometimes quite hard to detect. 
The `lag` function is a nice example of that. For ordinary vectors/data frames, `stats::lag` can lead to bugs but might be the default; meanwhile, `dplyr::lag` (or `data.table::shift(type = "lag")`) is clearer and more reliable.

```{r compare-lag-functions}
lag_example <- tibble::tibble(
  year = 2020:2022,
  gdp = c(100, 120, 140)
)

lag_example |>
  mutate(
    `dplyr::lag` = dplyr::lag(gdp),
    `stats::lag` = as.numeric(stats::lag(gdp))
  )
```

Similarly, if the code uses `select` without specifying that it should call `dplyr::select`, while `raster` is called, the code might break given the existence of `raster::select`. Here, at least, there wouldn't be a mysterious bug, unlike the `lag` case). 

Another example is `copy_labels`, which exists in both `sjlabelled` and `labelled`. The two versions do not use the same argument order, and they may also change values while copying labels, so it is worth writing the package name explicitly and checking the documentation.

# Data construction and cleaning

## Joins

Quite often we want to join datasets, using one or more common keys. Consider which rows should stay, how to treat units that do not match, whether the key is unique, whether the keys are consistent across datasets, and what to do when both datasets contain columns with the same name.

There are four common types of joins:

| Join | What it keeps | When to use |
|---|---|---|
| `left_join` | all rows from the first dataset | adding variables to a main dataset |
| `inner_join` | only rows that match in both datasets | keeping only units that exist in both sources |
| `full_join` | all rows from both datasets | checking coverage across sources, less common |
| `right_join` | all rows from the second dataset | same idea as `left_join`, but less convenient |

I usually like to have a "master" dataset with the units that are used for the final analysis, and then `left_join` other datasets into it. During data construction, `inner_join` and `full_join` can also be useful at times. 

Another consideration is the relationship between the keys:

| Relationship | Meaning | Example |
|---|---|---|
| one-to-one | each key appears once in both datasets | one row per country in both datasets |
| many-to-one | many rows in the main data match one row in the added data | country-year data joined to one region per country |
| one-to-many | one row in the main data matches many rows in the added data | one country joined to several yearly rows |
| many-to-many | keys repeat on both sides | usually risky unless explicitly intended |

The default in `dplyr` is  `relationship = NULL` which does not enforce a relationship, but gives a warning for many-to-many join that was not stated as such. If we know what the relationship should be, it can be useful to state it explicitly, for example `relationship = "many-to-one"`, so the code fails when the data do not match the assumption.

Another common issue is conflicting columns. If both datasets have `country` and `country_name`, for example, and we join on `country` only, then R may create columns such as `country_name.x` and `country_name.y`. That is sometimes useful for checking differences across sources, but usually it is cleaner to either use explicit suffixes or select away the unnecessary version before joining.

Assume that we have those two datasets, and we want to add the GDP variable into the country-region data:

```{r simple-country-join-example}
country_gdp <- tibble::tibble(
  iso3 = c("USA", "CHN", "IND"),
  country_name = c("United States", "China", "India"),
  gdp_b = c(23000, 18000, 3500)
)

country_regions <- tibble::tibble(
  iso3 = c("USA", "CHN", "IND", "MEX"),
  country_name = c("United States of America", "China", "India", "Mexico"),
  region = c("North America", "East Asia", "South Asia", "Latin America")
)

country_gdp

country_regions

```

```{r}

joined_country_regions <- country_regions |>
  left_join(country_gdp, join_by(iso3))

joined_country_regions

country_regions |>
  inner_join(country_gdp, join_by(iso3))

country_regions |>
  left_join(country_gdp, join_by(country_name))

country_regions |>
  left_join(
    country_gdp,
    join_by(iso3),
    suffix = c("_regions", "_gdp")
  )

country_regions |>
  left_join(
    country_gdp |> 
      select(-country_name),
    join_by(iso3)
  )
```

Here, both `left_join` and `inner_join` can add GDP, but they do not keep the same rows: `left_join` keeps Mexico and leaves GDP as `NA`, while `inner_join` drops Mexico. `full_join` and `right_join` also exist, but I use them less often.

In this example, joining by country name is problematic because the United States is written differently across the two datasets. The `country_name` columns also conflict after joining by `iso3`, so one option is to use suffixes and another is to drop the extra `country_name` column before joining.

## Messages and quick checks

One thing I have not done enough, but probably should, is adding a few short messages after important construction steps, to make mistakes easier to notice. Adding too many can make the code noisy, but after joins or aggregations they can be useful.

For example, after a join it is often worth printing how many units matched, how many did not, and how many rows have missing values in the new variable. After a `summarise`, one could print how many rows were collapsed or how many groups have missing values. 

```{r join-messages, message=TRUE}
joined_country_regions <- country_regions |>
  left_join(
    country_gdp |> 
      select(-country_name),
    join_by(iso3)
  )

unmatched_country_regions <- country_regions |>
  filter_out(iso3 %in% country_gdp$iso3)

message(
  n_distinct(country_regions$iso3) - nrow(unmatched_country_regions), " of ",
  n_distinct(country_regions$iso3), " country-region units matched to GDP."
)

message(
  nrow(unmatched_country_regions), " country-region units did not match to GDP."
)

message(
  sum(is.na(joined_country_regions$gdp_b)), " rows have missing GDP after the join."
)
```

## Aggregates

In some datasets there may be aggregates (e.g., "Middle East" in a country-level dataset). In some datasets there are variables that refer to those, but not always, and the names of the 'aggregate' countries might be different across different datasets (e.g., in two different datasets of FAOSTAT, there is in one dataset (https://www.fao.org/faostat/en/#data/FBSH) an aggregated region of 'Low Income Food Deficit Countries', but in another (https://www.fao.org/faostat/en/#data/QCL) this appears as 'Low Income Food Deficit Countries (LIFDCs)'. Given that there is not always an ISO2/3 equivalent for each aggregate, one needs to pay attention to those cases. 

<!-- To add: perhaps turn this into a short checklist: identify aggregate rows; decide whether to keep, drop, or reconstruct them; document the choice; and compare constructed aggregates to a reference source when possible. -->

<!-- What if we want to aggregate multiple rows. I've removed before the EU as an aggregate, but let's assume that we'd want to construct it from scratch: -->
<!-- ```{r} -->
<!-- eu_df <- country_panel |>  -->
<!--   mutate( -->
<!--     iso2 = if_else(iso2 %in% eu_iso2, "EU", iso2), -->
<!--     country_name = if_else(iso2 == "EU", "European Union", country_name) -->
<!--   ) |> -->
<!--   summarise( -->
<!--     gdp_b = sum(gdp_b, na.rm = TRUE), -->
<!--     pop_m = sum(pop_m, na.rm = TRUE), -->
<!--     .by = c(iso2, country_name, year) -->
<!--   ) -->
<!-- ``` -->

<!-- # eu_comparison <- eu_df |>  -->
<!-- #   filter(iso2 == "EU") |>  -->
<!-- #   select(year, eu_constructed_gdp_b = gdp_b, eu_constructed_pop_m = pop_m) |> -->
<!-- #   left_join( -->
<!-- #     country_panel_all |>  -->
<!-- #       filter(iso3 == "EUU") |>  -->
<!-- #       select(year, eu_wdi_gdp_b = gdp_b, eu_wdi_pop_m = pop_m), -->
<!-- #     by = "year" -->
<!-- #   ) |> -->
<!-- #   mutate( -->
<!-- #     gdp_diff_b = round(eu_constructed_gdp_b - eu_wdi_gdp_b, 2), -->
<!-- #     pop_diff_m = eu_constructed_pop_m - eu_wdi_pop_m -->
<!-- #   ) -->
<!-- #  -->
<!-- # eu_comparison -->

<!-- Importantly for cases like this, ideally we'd want to verify that the values make sense compared to some reference data. For instance, I would try to find data on the aggregate (EU, in this case) from another source - IMF, Eurostat, etc, asking an AI whether it looks plausible etc.  -->

# Country level analysis

A lot of the analysis I have done was at the country level, or used countries as parent units. Thus, most of this text is going to use countries as example units, but here are some tips on dealing with country level data:

## Country identifiers
Country names are useful for plots and tables, but they are often inconsistent across datasets. For example, "Ivory Coast" and "Cote d'Ivoire" may refer to the same country. For joins, it is usually better to rely on ISO codes, while keeping the names as labels.

Even country codes need a quick check. For example:
1) South Sudan's ISO3 is usually "SSD", but some datasets may use "SDS"
2) Namibia's ISO2 is "NA", which can be confused with a missing value if the data is read carelessly

Normally the WDI includes data on aggregates (if you download directly with 'WDI::WDI(), use extra = TRUE')). For illustration, however, I have downloaded the data from the WDI website as csv. https://data.worldbank.org/indicator/SP.POP.TOTL and https://data.worldbank.org/indicator/NY.GDP.MKTP.KD

```{r build-country-panel}
pop <- read_csv(here("data", "API_SP.POP.TOTL_DS2_en_csv_v2_33112.csv"), skip = 4) |>
  pivot_longer(
    cols = matches("^\\d{4}$"),
    names_to = "year",
    values_to = "pop"
  ) |>
  transmute(
    iso3 = `Country Code`,
    country_name = `Country Name`,
    year = as.integer(year),
    pop
  )

gdp <- read_csv(here("data", "API_NY.GDP.MKTP.KD_DS2_en_csv_v2_34247.csv"), skip = 4) |>
  pivot_longer(
    cols = matches("^\\d{4}$"),
    names_to = "year",
    values_to = "gdp"
  ) |>
  transmute(
    iso3 = `Country Code`,
    country_name = `Country Name`,
    year = as.integer(year),
    gdp
  )

country_panel_all <- full_join(pop, gdp, by = c("iso3", "country_name", "year")) |>
  transmute(
    iso2 = countrycode::countrycode(iso3, "iso3c", "iso2c", warn = FALSE),
    iso3,
    country_name,
    year,
    gdp_b = round(gdp / 1e9, 5),
    pop_m = round(pop / 1e6, 3)
  ) |>
  filter(year >= 2000) |>
  arrange(iso2, year)

head(country_panel_all)
```

This WDI file also includes rows that are not countries, such as regions or income groups. One quick diagnostic is to inspect the ISO3 codes that failed to convert to ISO2. Most failed conversions are aggregates, but this check also catches real places that need manual handling, like Kosovo.

```{r list-unmatched-countries}
unmatched_iso3 <- country_panel_all |> 
  filter(is.na(iso2)) |> 
  distinct(iso3, country_name) |> 
  arrange(country_name)

tibble::tibble(n_unmatched_iso3 = nrow(unmatched_iso3))

kosovo_row <- which(unmatched_iso3$iso3 == "XKX")

unmatched_iso3 |> 
  slice(max(1, kosovo_row - 2):min(n(), kosovo_row + 2))
```

Thus, I'd manually add an ISO2 code for Kosovo. I also keep the EU as a custom aggregate because I want to use it later.

```{r fix-country-codes}
country_panel <- country_panel_all |> 
  mutate(
    iso2 = case_when(
      iso3 == "XKX" ~ "XK",
      iso3 == "EUU" ~ "EU", 
      TRUE ~ iso2
      )
  ) |> 
  filter(!is.na(iso2))
```

## Geographies consistency over time

Various geographical units of observations do not have consistent boundaries over time. Countries change their borders, gain independence or split; admin units sometimes change have changed their borders. Different datasets treat those changes differently - perhaps a value in a given year refers to the current border or the old one. 
When working with country-level data, for example, pay attention to countries whose borders have changed in the period of analysis (or recently!). 'Sudan' might reflect the current state of Sudan, or also the values of South Sudan. Montenegro might be part of Yugoslavia, Serbia and Montenegro, or Montenegro; the same is true for Kosovo (despite not being mentioned in the name of 'Serbia and Montenegro'), although Kosovo might not even exist in certain datasets. Taiwan might be referred to in different ways, depending on the dataset.

<!-- To add: perhaps make this a checklist too: which boundaries does the dataset use; are units historical or current; were any units split/merged during the period; are successor states treated consistently; and do plots around the transition years reveal suspicious jumps. -->

This, for example, looks rather fine - there are no obvious spikes in the data around the year of independence:
```{r plot-sudan-south-sudan}
country_panel |>
  filter(iso2 %in% c("SD", "SS"), year %in% 2006:2014) |>
  ggplot() + 
  geom_line(aes(x = year, y = pop_m, group = country_name, color = country_name))

```

# Missing values
Assume that we have a dataset with some missing values:
```{r make-na-example}
na_df <- country_panel |>
  filter(
    iso2 %in% c("CU", "ER", "BO"), 
    year %in% 2023:2025
    ) |>
  select(year, country_name, gdp_b) |> 
  pivot_wider(
    names_from = country_name,
    values_from = c(gdp_b)
  )

na_df
```
We want to aggregate that data. The main options are 1) na.rm = TRUE, which removes the NAs before calculating, such that sum(df$x, na.rm = TRUE) would still return a value. For sums this can look like treating NA as 0, and if all values are NA it returns 0, but for means it also changes the denominator; 2) na.rm = FALSE, which forces the aggregation to have NA if any of the values is NA 

Especially if we use a loop or across to aggregate over several variables or units, it can be a problem
Say that we want to sum GDP, in billions, for each of those three countries in this period.  
```{r compare-na-sums}
na_df_long <- na_df |> 
  pivot_longer(
    -year,
    names_to = "country_name",
    values_to = "gdp_b"
  )

na_rm_options <- c("na.rm = TRUE", "na.rm = FALSE", "NA only if all missing")

bind_rows(
  na_df_long |> 
    summarise(
      method = "na.rm = TRUE",
      sum = sum(gdp_b, na.rm = TRUE),
      .by = country_name
    ),
  na_df_long |> 
    summarise(
      method = "na.rm = FALSE",
      sum = sum(gdp_b, na.rm = FALSE),
      .by = country_name
    ),
  na_df_long |> 
    summarise(
      method = "NA only if all missing",
      sum = if (all(is.na(gdp_b))) NA_real_ else sum(gdp_b, na.rm = TRUE),
      .by = country_name
    )
  ) |> 
  arrange(country_name, match(method, na_rm_options)) |> 
  pivot_wider(
    names_from = method,
    values_from = sum
  ) |> 
  left_join(
    na_df_long |> 
      summarise(
        n_observed = sum(!is.na(gdp_b)),
        .by = country_name
      ),
    by = "country_name"
  )
```

so the table lets us see the three choices together: when we use na.rm = FALSE, a column that has values and NAs is always NA. But perhaps we are interested in preserving the NAs only in cases where all rows are NA. then, we need to modify the function. It is also often useful to show the amount of missingness itself. Otherwise, two columns can have the same sum while one of them is based on much less data.

# Plotting

Plots are great. I mainly use them for two, somewhat different purposes - inspecting the data and presenting results. Graphs can highlight weird data spikes, missing or implausible values, certain units driving the results in a (much) clearer way than summary tables. 
When presenting results, we want the readers to understand the main message, ideally without working too hard.
To me, graphs in economic papers tend to be quite bad in serving the latter goal; they're often ugly, too busy and hard to understand (whether it's due to lack of information, to low priorities to visualiations, to standartization of those types of graphs or anything else is a differnet question). That is sometimes necessary, especially for readers who want to inspect the details, but often the graph becomes less useful because of it.

There are much better guides than whatever I will write here. I really like [Saloni Dattani's guide](https://www.scientificdiscovery.dev/p/salonis-guide-to-data-visualization).

One should consider which type of graph would be useful. I think that line plots are suitable for most cases, especially for changes over time, especially with a small number of series or when some series need to be highlighted. Bar plots can work well for comparing categories, if there are not too many of them; while stacked bars (or areas) can be nice at times but can be hard to read unless the question is really about the total, the first category in the stack, or categories that undertook massive changes. Pie charts should be avoided for similar reasons (but much harder to read)

The [R Graph Gallery](https://r-graph-gallery.com/) is also a nice source for examples.

Some things to consider:

Axis titles, especially y axis titles, are often unnecessary, hard to read, and distracting. The same information can usually be conveyed using the main plot title. Very often, the x axis indicates time, which means that indicating the unit is unnecessary, though it still appears on many graphs in economics papers.

```{r plot-gdp-pc-raw}
eu27_countries <- c(
  "AT", "BE", "BG", "HR", "CY", "CZ", "DK", "EE", "FI", "FR",
  "DE", "GR", "HU", "IE", "IT", "LV", "LT", "LU", "MT", "NL",
  "PL", "PT", "RO", "SK", "SI", "ES", "SE"
)
highlight_countries <- c("IE", "LU", "SE")

plot_df <- country_panel |> 
  filter(
    iso2 %in% eu27_countries,
    year >= 2000,
    year <= 2024
  ) |> 
  mutate(
    gdp_pc = gdp_b * 1000 / pop_m,
    highlighted = if_else(iso2 %in% highlight_countries, iso2, "Other countries")
  )

plot_df |> 
  ggplot() + 
  geom_line(aes(x = year, y = gdp_pc, color = iso2)) +
  scale_x_continuous(breaks = seq(2000, 2024, 4), limits = c(2000, 2024)) +
  scale_y_continuous(limits = c(0, NA), labels = label_dollar())
```
Very minimal adjustments make it a lot cleaner:
```{r plot-gdp-pc-clean}
plot_df |> 
  ggplot() + 
  geom_line(aes(x = year, y = gdp_pc, color = iso2)) + 
  scale_x_continuous(breaks = seq(2000, 2024, 4), limits = c(2000, 2024)) +
  scale_y_continuous(limits = c(0, NA), labels = label_dollar()) +
  theme_minimal() + 
  theme(axis.title = element_blank()) + 
  labs(title = "GDP per capita across EU countries")
```

The legend can also often be avoided by placing the labels directly near the lines. Wouldn't always work (generally useful if there aren't many categories), but it's sometimes useful. The plot can also benefit from clearer colours (that are more colour-blind friendly). Say that we want to focus on Ireland's rise from Sweden's level of GDP per capita to nearly Luxembourg's, then these could be higlighted with the other EU countries having grey :

```{r plot-gdp-pc-highlighted}
plot_colors <- c(
  "IE" = "#0072B2",
  "LU" = "#D55E00",
  "SE" = "#009E73",
  "Other countries" = "#b8bec6"
)

line_labels <- plot_df |> 
  filter(iso2 %in% highlight_countries) |> 
  summarise(
    gdp_pc = gdp_pc[which.max(year)],
    year = max(year),
    label = country_name[which.max(year)],
    .by = iso2
  )

plot_df |> 
  ggplot(aes(x = year, y = gdp_pc, group = iso2)) + 
  geom_line(aes(color = highlighted), linewidth = 0.9, alpha = 0.85) + 
  geom_text(
    data = line_labels,
    aes(label = iso2, color = iso2),
    hjust = 0,
    nudge_x = 0.25,
    show.legend = FALSE
  ) +
  scale_color_manual(values = plot_colors) +
  scale_x_continuous(breaks = seq(2000, 2024, 4), limits = c(2000, 2025.5)) +
  scale_y_continuous(limits = c(0, NA), labels = label_dollar()) +
  theme_minimal() + 
  theme(
    axis.title = element_blank(),
    legend.position = "none",
    plot.margin = margin(5.5, 60, 5.5, 5.5)
  ) + 
  coord_cartesian(clip = "off") +
  labs(title = "Ireland spiked toward the EU's richest")
```

Finally, a more polished version could use background colours, cleaner axis, using the actual names as labels and so on:

```{r plot-gdp-pc-final}
plot_df |> 
  ggplot(aes(x = year, y = gdp_pc, group = iso2)) + 
  geom_line(aes(color = highlighted), linewidth = 0.9, alpha = 0.8) + 
  geom_text(
    data = line_labels,
    aes(label = label, color = iso2),
    hjust = 0,
    nudge_x = 0.25,
    fontface = "bold",
    show.legend = FALSE
  ) +
  scale_color_manual(values = plot_colors) +
  scale_x_continuous(breaks = seq(2000, 2024, 4), limits = c(2000, 2025.5)) +
  scale_y_continuous(limits = c(0, NA), labels = label_dollar()) +
  labs(
    title = "Ireland spiked toward the EU's richest",
    subtitle = "GDP per capita, constant 2015 US dollars",
    caption = "Source: The World Bank"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.title = element_blank(),
    legend.position = "none",
    plot.background = element_rect(fill = "#f7f8f3", color = NA),
    panel.background = element_rect(fill = "#f7f8f3", color = NA),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    plot.caption = element_text(color = "#6b7280"),
    plot.margin = margin(5.5, 70, 5.5, 5.5)
  ) +
  coord_cartesian(clip = "off")
```

Another useful AI trick is to ask it to follow a specific visualization guide rather than asking for a "better plot" in general. For example, one could ask for suggestions inspired by the *Financial Times* visual vocabulary, the Urban Institute data visualization style guide, or the Datawrapper/Flourish style of direct labeling and restrained annotation.

# Spatial Analysis

Quite a lot of the analysis I have done is related to spatial data, so I include a few practical notes here.

There are two main types of spatial datasets. Vector data includes discrete geometries, such as points, lines, and polygons. Raster data stores values in grid cells. 

Note that vector data is often saved as shapefiles. Shapefiles are split across several files (`.shp`, `.shx`, `.dbf`, etc.), have short column-name limits, do not handle mixed geometry types well, and are not especially efficient. **`gpkg` is often preferable**: it is one file, handles longer column names, can store multiple layers, and is usually easier to move around.

**For raster data, I would usually use `terra`** in R. Raster data is often saved as `.tif`, especially for ordinary gridded layers. Another common format is `.nc` or NetCDF, which is often used for climate or other data with several dimensions. In R, I would usually try `terra::rast()` first for both `.tif` and `.nc` files. If the NetCDF file has many variables or time layers, it is worth checking the layer names after reading it and selecting the variable or band we actually need.

**For vector data, I would usually use `sf`** in R. `sf::read_sf()` and `sf::write_sf()` are usually convenient wrappers around `st_read()` and `st_write()`, so I would use them most of the time (especially when reading a file with one obvious layer). 

## Which tool to use

For me, there are mainly three types of tools that are useful in spatial analysis.

| Type | Examples | Main advantages | Main drawbacks |
|---|---|---|---|
| Programming languages | R with `sf` and `terra`; Python with `geopandas`, `rasterio`, `xarray`, etc. | Reproducible workflows; easy to combine spatial work with data cleaning, modeling, and plotting; good for local vector/raster operations | Local storage and memory can become a problem; worse for some visual validation |
| GIS software | QGIS, ArcGIS | Inspection, validation, manual exploration, and cartographic output | requires local storage; less convenient for some analysis workflows |
| Google Earth Engine | GEE Code Editor, GEE Python API | Hosted data catalogs; convenient raster and remote-sensing operations; interactive map checking; avoids many local storage problems | Less intuitive coding and debugging; quotas and server-side limits; less convenient for some reproducible project workflows |

**Programming languages** - can do most of the work well. Running things is rather fast for most cases and visualizations (e.g., `ggplot2::geom_sf()`) are decent. Mainly, they are very convenient when spatial work is a part of a larger data construction or analysis workflow. 

**GIS software** - ArcGIS is powerful but not free; QGIS is very good for visualizations and debugging or validating data (e.g., examine whether a spatial join worked properly or matched the wrong places). It is also possible to program there, but I find the programming less convenient than `sf` in R (decide for yourself, though; perhaps starting out with QGIS is easier), and it creates an additional layer of workflow (save the data, load it elsewhere, etc.). I personally like working with R though, especially with AI agents helping write and debug code. 

**GEE** - useful mainly, in my opinion, when the data already exists in the Earth Engine data catalog (https://developers.google.com/earth-engine/datasets) or the GEE community catalog (https://gee-community-catalog.org/). For example, using Hansen forest loss data or built-up data from GHSL is easier when one can avoid downloading and storing the data, and instead we'd run the relevant operations there. It can also be advantageous for raster or remote-sensing operations, such as filtering large image collections by date/cloud cover/location, combining multiple images, and using reducers over space or time. However, this is less of a reason to use GEE for every heavy spatial process now that there are quotas for noncommercial users. It is also good for validation and checking layers on a map. The main drawback of GEE, beyond quotas, is the coding experience: the Code Editor uses the Earth Engine JavaScript API, many objects are server-side, and errors can be opaque. GEE now has a Gemini assistant in the Code Editor (which I haven't tried yet), which requires a Gemini API key, so it is not necessarily unlimited/free. In terms of workflows, it can be annoying, although it's possible to use the Python API to keep all the code in a single project with other R/Python/Stata codes.

When unsure which tool to use, AIs would probably be useful to help you judge, and now the costs of writing a sample code in R and examining its efficiency are so low that it's probably worth starting from it.

### Recommendations for working with GEE

If there is one set of units that will be used across multiple scripts, I would usually create a master feature collection and call it at the start of each script, for example `var worldUnits = ee.FeatureCollection('projects/ee-your_project/assets/worldUnits');`. Then, we can have a master script that loads the unit file and then calls separate modules.

I would pay attention to boundary datasets that AIs tend to suggest, such as `FAO/GAUL/2015/level0`. They can be useful, but if we later group by ISO3 or compare to country-level data, some territories or administrative units may need to be merged first. In those cases it can be better to have a separate GEE script that cleans the boundary file, matches names (e.g., to ISO3), merges territories that should belong to the same unit, and exports the result as a reusable GEE asset. Then the analysis scripts can call that asset.

## Spatial matching

A unit may intersect several polygons, a point may fall just outside a boundary because the datasets use slightly different borders, or a raster cell may only partly overlap the polygon. In those cases we need to decide whether to keep all matches, use the largest overlap, weight by area or population, use the nearest feature, or drop ambiguous cases. Whatever choice is made, it is worth saving a few examples and plotting them.

I think it is useful to separate the cases, because the main mistakes differ:

| Match type | Common tools | Main question |
|---|---|---|
| point to polygon | `st_join(..., join = st_within)` | which unit contains each point? | 
| polygon to polygon | `st_intersection()` | which areas overlap, and by how much? |
| nearest vector match | `st_nearest_feature()`, `nngeo::st_nn()` | which feature is closest when there is no exact match? |
| raster to polygon | `terra::extract()` | how should grid-cell values be summarized within each polygon? |

For point-to-polygon matching, I would usually use `st_join(points, polygons, join = st_within)`, which attaches the polygon attributes to each point that falls inside a polygon. After this, I would check unmatched points.

For polygon-to-polygon matching, note that there are two `sf` functions with similar names: `st_intersects()` and `st_intersection()`. `st_intersection()` returns a new spatial dataset whose rows are the overlapping pieces. For example, if one country overlaps three ecological zones, the result can have three rows for that country, each with a smaller geometry and attributes from both the country layer and the ecological-zone layer. This is useful when we need to calculate overlap areas or area-weighted variables. `st_intersects()` only tells us which geometries touch or overlap, so it can be useful as a quicker diagnostic, or as a first filtering step before running a heavier `st_intersection()`.

Nearest-feature matching can be useful for vector data when there is no exact match. `st_nearest_feature()` gives the single nearest feature, while `nngeo::st_nn()` is useful when we want several nearest neighbors, or when we only want matches within a given distance. For example, we might match a mine to the nearest road, but only if the road is within 10 km.

For raster-to-polygon matching, the question is usually how to summarize cells inside each polygon. For example, if we have a population raster and administrative units, we may want total population in each unit. If we have a temperature raster, we may want the mean. 
Pay attention to partial overlaps, as a cell may sit partly inside a polygon and partly outside it. In `terra::extract()`, the default does not return cell fractions; it extracts the selected raster cells and applies the function we ask for, such as `mean` or `sum`. Setting `weights = TRUE` returns approximate fractions of cells covered by the polygon, while `exact = TRUE` is more precise but slower. Those fractions are not always the right choice, but they can be useful. For example, for population we might sum `population * fraction`, while for temperature we might calculate a weighted mean.

## creating visualizations

For maps in R, `ggplot2` with `geom_sf()` is usually enough. A choropleth map usually needs two inputs: boundaries and a variable to color each area.

The example below is useful because it first highlights a bug. If we join by ISO2, France is not matched because the `rnaturalearth` file uses `-99` as its ISO2 code. That is exactly the kind of problem maps can help catch quickly.
```{r map-eu-gdp-pc, message=FALSE}
eu_gdp_2024 <- country_panel |>
  filter(year == 2024, iso2 %in% eu27_countries) |>
  mutate(gdp_pc = gdp_b * 1000 / pop_m) |>
  select(iso2, iso3, gdp_pc)

europe_boundaries <- rnaturalearth::ne_countries(
  scale = "small",
  returnclass = "sf"
) |> 
  select(iso2 = iso_a2, iso3 = adm0_a3, continent, geometry) |>
  filter(continent == "Europe")

bad_iso2_map <- europe_boundaries |>
  mutate(in_eu = iso2 %in% eu27_countries) |>
  left_join(
    eu_gdp_2024 |> select(iso2, gdp_pc),
    join_by(iso2)
  )

bad_iso2_map |>
  filter(iso3 %in% c("FRA", "DEU", "ESP")) |>
  st_drop_geometry() |>
  select(iso2_boundary = iso2, iso3, in_eu, gdp_pc)

bad_iso2_map |>
  ggplot() +
  geom_sf(fill = "#d9d9d9", color = "white", linewidth = 0.12) +
  geom_sf(
    data = bad_iso2_map |> filter(in_eu),
    aes(fill = gdp_pc),
    color = "white",
    linewidth = 0.12
  ) +
  scale_fill_viridis_c(
    labels = label_number(scale = 1 / 1000, suffix = "k"),
    na.value = "#eeeeee",
    guide = guide_colorbar(
      title = NULL,
      barwidth = grid::unit(160, "pt"),
      barheight = grid::unit(8, "pt"),
      label.position = "bottom"
    )
  ) +
  coord_sf(xlim = c(-12, 35), ylim = c(34, 72), expand = FALSE) +
  theme_void() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    plot.caption = element_text(color = "#6b7280")
  ) +
  labs(
    title = "GDP per capita across EU countries",
    subtitle = "Reveals a bad join: France is not merged",
    caption = "Source: The World Bank"
  )
```

The fix is to join by ISO3. For the final display, I would also usually consider binning the values rather than using a continuous color scale, with the bins depending on the comparison we want to make. The first version below keeps a wide top category, which is useful for separating poorer, middle, rich, and very rich countries. The second version uses the same palette but compresses the top and gives more detail in the middle and bottom. 
```{r map-eu-gdp-pc-fixed, message=FALSE}
eu_gdp_map <- europe_boundaries |>
  mutate(in_eu = iso3 %in% eu_gdp_2024$iso3) |>
  left_join(
    eu_gdp_2024 |> select(iso3, gdp_pc),
    join_by(iso3)
  ) |>
  mutate(
    gdp_pc_group_wide = cut(
      gdp_pc,
      breaks = c(0, 20000, 35000, 50000, 70000, Inf),
      labels = c("<$20k", "$20k-$35k", "$35k-$50k", "$50k-$70k", "$70k+"),
      right = FALSE
    ),
    gdp_pc_group_middle = cut(
      gdp_pc,
      breaks = c(0, 10000, 15000, 20000, 30000, 40000, Inf),
      labels = c("<$10k", "$10k-$15k", "$15k-$20k", "$20k-$30k", "$30k-$40k", "$40k+"),
      right = FALSE
    )
  )

eu_gdp_map |>
  ggplot() +
  geom_sf(fill = "#d9d9d9", color = "white", linewidth = 0.12) +
  geom_sf(
    data = eu_gdp_map |> filter(in_eu),
    aes(fill = gdp_pc_group_wide),
    color = "white",
    linewidth = 0.12
  ) +
  scale_fill_manual(
    values = c(
      "<$20k" = "#edf8fb",
      "$20k-$35k" = "#b2e2e2",
      "$35k-$50k" = "#66c2a4",
      "$50k-$70k" = "#2ca25f",
      "$70k+" = "#006d2c"
    ),
    na.value = "#eeeeee",
    name = NULL
  ) +
  coord_sf(xlim = c(-12, 35), ylim = c(34, 72), expand = FALSE) +
  theme_void() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    plot.caption = element_text(color = "#6b7280")
  ) +
  labs(
    title = "GDP per capita across EU countries",
    subtitle = "Constant 2015 US dollars, not PPP-adjusted",
    caption = "Source: The World Bank"
  )

eu_gdp_map |>
  ggplot() +
  geom_sf(fill = "#d9d9d9", color = "white", linewidth = 0.12) +
  geom_sf(
    data = eu_gdp_map |> filter(in_eu),
    aes(fill = gdp_pc_group_middle),
    color = "white",
    linewidth = 0.12
  ) +
  scale_fill_manual(
    values = c(
      "<$10k" = "#edf8fb",
      "$10k-$15k" = "#ccece6",
      "$15k-$20k" = "#99d8c9",
      "$20k-$30k" = "#66c2a4",
      "$30k-$40k" = "#2ca25f",
      "$40k+" = "#006d2c"
    ),
    na.value = "#eeeeee",
    name = NULL
  ) +
  coord_sf(xlim = c(-12, 35), ylim = c(34, 72), expand = FALSE) +
  theme_void() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    plot.caption = element_text(color = "#6b7280")
  ) +
  labs(
    title = "GDP per capita across EU countries",
    subtitle = "Same palette, narrower bins",
    caption = "Source: The World Bank"
  )
```

## finding representative points of a polygons
sometimes we may want to work with single points in a given unit, rather than the full polygon. that's especially the case if 1) the analysis is very heavy (e.g., calculating road path from unit x to y); 2) we are interested in the role of specific areas within the unit (e.g., the main city), 3) fill
if we chooes to use points, how do we select the point? it depends obviously on what we want to achieve.
centroid is a popular solution, but it suffers from various drawbacks. 
- it is not necessarily inside the unit. think of units with weird shapes (vietnam), that are composed of many distinct geometries (multiple-island countries, like the philippines) or with holes inside (south africa). if we're calculating the distance of centroid of country x to that of y, and country y's centroid is inside that of x's, it's not ideal...
- one solution for that would be to use `st_point_on_surface` from the sf package. it guarantees that the point would be on the surface of the unit, and generally is around its middle. however, for certain cases, it might still not capture something we're interested in. I was working with an admin unit that included a city in the mainland, and an island very far from the mainland. The island was larger, so `st_point_on_surface` has led to a point in the island. In some cases perhaps it's fine, but if we're interested in anything that involves population - then it really does not capture what we want

```{r plot-centroid-surface-points, echo=FALSE}
normal_polygon <- st_polygon(list(
  matrix(
    c(
      0.2, 0.1,
      4.1, 0.0,
      4.4, 2.2,
      3.6, 4.0,
      1.3, 4.3,
      0.0, 2.5,
      0.2, 0.1
    ),
    ncol = 2,
    byrow = TRUE
  )
))

hole_polygon <- st_polygon(list(
  matrix(c(0, 0, 4, 0, 4, 4, 0, 4, 0, 0), ncol = 2, byrow = TRUE),
  matrix(c(1.2, 1.2, 1.2, 2.8, 2.8, 2.8, 2.8, 1.2, 1.2, 1.2), ncol = 2, byrow = TRUE)
))

schematic_units <- st_sf(
  case = c("compact polygon", "polygon with hole"),
  geometry = st_sfc(normal_polygon, hole_polygon)
)

centroids <- schematic_units |>
  st_centroid() |>
  mutate(point_type = "centroid")

surface_points <- schematic_units |>
  st_point_on_surface() |>
  mutate(point_type = "point on surface")

bind_rows(centroids, surface_points) |>
  ggplot() +
  geom_sf(data = schematic_units, fill = "#e5e7eb", color = "#4b5563") +
  geom_sf(aes(color = point_type), size = 2.8) +
  facet_wrap(~case) +
  scale_color_manual(values = c("centroid" = "#D55E00", "point on surface" = "#0072B2")) +
  coord_sf(datum = NA) +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.title = element_blank()
  )
```

- what about population-weighted centroid? we can use data like GHS-POP to do it (recommended to do it in GEE). However, 1) it might still be outside of the unit, 2) say that there are several population centres in different areas. the pop-weighted centroid is probably going to be between them, which might be good but it'll likely reflect a (relatively) remote point. which might be fine but not always!
```{r plot-population-centroid-points, echo=FALSE, fig.width=6, fig.height=4}
population_unit <- st_sf(
  geometry = st_sfc(normal_polygon)
)

mock_population_grid <- st_sf(
  geometry = st_make_grid(population_unit, n = c(10, 10))
) |>
  st_intersection(population_unit)

grid_centers <- st_coordinates(st_centroid(mock_population_grid))

mock_population_grid <- mock_population_grid |>
  mutate(
    x_coord = grid_centers[, 1],
    y_coord = grid_centers[, 2],
    population = round(
      15 +
        950 * exp(-((x_coord - 1.0)^2 + (y_coord - 3.0)^2) / 0.18) +
        240 * exp(-((x_coord - 3.2)^2 + (y_coord - 1.0)^2) / 0.14) +
        180 * exp(-((x_coord - 3.4)^2 + (y_coord - 3.1)^2) / 0.16) +
        130 * exp(-((x_coord - 2.0)^2 + (y_coord - 1.9)^2) / 0.12)
    )
  )

population_weighted_point <- mock_population_grid |>
  st_drop_geometry() |>
  summarise(
    x_coord = weighted.mean(x_coord, population),
    y_coord = weighted.mean(y_coord, population)
  ) |>
  st_as_sf(coords = c("x_coord", "y_coord")) |>
  mutate(point_type = "population-weighted centroid")

highest_population_point <- mock_population_grid |>
  slice_max(population, n = 1, with_ties = FALSE) |>
  st_centroid() |>
  mutate(point_type = "highest population cell")

population_centroid <- population_unit |>
  st_centroid() |>
  mutate(point_type = "centroid")

population_surface_point <- population_unit |>
  st_point_on_surface() |>
  mutate(point_type = "point on surface")

population_summary_points <- bind_rows(
  population_centroid,
  population_surface_point,
  population_weighted_point,
  highest_population_point
)

population_summary_labels <- population_summary_points |>
  mutate(
    point_x = st_coordinates(geometry)[, 1],
    point_y = st_coordinates(geometry)[, 2],
    label_x = case_when(
      point_type == "highest population cell" ~ point_x - 0.16,
      point_type == "population-weighted centroid" ~ point_x + 0.18,
      point_type == "centroid" ~ point_x + 0.18,
      TRUE ~ point_x - 0.16
    ),
    label_y = case_when(
      point_type == "highest population cell" ~ point_y + 0.20,
      point_type == "population-weighted centroid" ~ point_y - 0.18,
      point_type == "centroid" ~ point_y + 0.18,
      TRUE ~ point_y - 0.20
    )
  ) |>
  st_drop_geometry()

ggplot() +
  geom_sf(
    data = mock_population_grid,
    aes(fill = population),
    color = "white",
    linewidth = 0.2
  ) +
  geom_sf(data = population_unit, fill = NA, color = "#4b5563", linewidth = 0.5) +
  geom_sf(
    data = population_summary_points,
    aes(color = point_type),
    size = 3
  ) +
  geom_label(
    data = population_summary_labels,
    aes(x = label_x, y = label_y, label = point_type),
    size = 3.1,
    linewidth = 0.15,
    label.padding = unit(0.08, "lines")
  ) +
  scale_fill_gradient(low = "white", high = "#b91c1c") +
  scale_color_manual(
    values = c(
      "centroid" = "#D55E00",
      "point on surface" = "#0072B2",
      "population-weighted centroid" = "#009E73",
      "highest population cell" = "#00BFC4"
    )
  ) +
  coord_sf(datum = NA, expand = TRUE) +
  theme_minimal(base_size = 12) +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  )
```

- similarly to the last suggestion, can use 'the most populated point'. for certain things it might be preferable (for others, not as much). one thing to consider here, as many spatial analysis, is the resolution. the most populated 100m place in a country could theoretically be a small town that has an extremely dense point. so aggregating by 1/5/10km might be preferable, but examine the outputs and see whether this captures what you're interested in

### finding a 'representative' category
there are various spatial datasets that might involve finding some balance between various categories. mines datasets, for example, sometimes have each individual mine as a row, or would have a value of 'suitability for a given commodity' per cell. depends on what we want to have, we might want to aggregate the data per unit. should we sum the number of distinct mines? use the commodity with the highest value of suitability? count whether any mine of each type exists? There is no obvious answer here. The right choice depends on whether the variable is meant to capture abundance, dominance, exposure, or just the presence of something.

# Analysis

small note - if using `log()`, check the values and units first. Logging GDP in dollars is not the same as logging GDP in millions, and logging very small values can produce numbers that are technically correct but not very meaningful. 

## shift share

to be added

# Debugging and validating results

- row counts before and after each important step
- duplicate keys
- unmatched observations
- impossible values
- missingness summaries
- graphs
- edge cases
- compare aggregates to an external source
- inspect examples rather than only summary statistics

# constructing workflows

I like organising both code and data in raw, intermediate/build, and final/output folders. raw data should usually be kept as close as possible to the original files, intermediate data should be the products of construction steps, and final/output data should be the version that is actually used for analysis, tables, or figures. This makes it easier to know what can be recreated and what should not be manually edited.

For code, I usually prefer numbered scripts or folders when there is a logical order, for example `01_import`, `02_clean`, `03_construct`, `04_analysis`. 
If some steps are expensive, it is often worth saving intermediate outputs and reusing them, but I would name them clearly and avoid keeping several unexplained versions of the same object.

Version control (usually git) is extremely useful. In projects with many authors, it can be useful to deal with conflicts; but even for a project that only you work on (or just you update the git) it is recommended, for tracking changes over time and reverting them more easily. For (especially heavy) data files, outputs that are repeatedly generated with different names (e.g., results_\{date\}) or private files, should go in `.gitignore` and not be tracked. 

For reproducibility, I would keep the seed discussion near the code that actually uses randomness. `set.seed(123)` is deterministic, like any other fixed seed; it is not more random because the number is familiar. If there is no reason to use a specific seed, one option is to randomly generate a seed once, record it, and then keep using that same seed throughout the project.

Before sending results, I would rerun the relevant scripts from a clean session and check the main outputs. At minimum, this usually means row counts, duplicate keys, unmatched observations, summary statistics, missing values, and zeros in places where zeros may indicate a construction problem.

## Handoff notes

<!-- To add later. -->

## Definitions and units

<!-- To add later. -->

## Assumptions and judgment calls

<!-- To add later. -->

# AI use

This would probably change quite a bit soon, but the setup that I like now for working with AI agents is to separate reusable preferences from project-specific context. One possible structure is:

```text
ai_instructions/
  general/
    coding_style.md
    data_checks.md
    writing_style.md
  project/
    context.md
    pipeline.md
    datasets.md
```

The general files can include coding conventions, for example using `|>` rather than `%>%`, and recurring data-construction checks, such as looking for missing values after merges or joins that multiply rows. The project files can describe the question, the main unit of observation, the code structure, and the intended order of the data pipeline. This does use context, so I would not overdo it for small tasks, but for longer projects it can save repeated explanations and reduce mistakes. If there are project-specific coding conventions, I would usually add them as short additions to the general convention rather than copying the whole thing again.

There are also many good skills online, so there is probably no need to create many new ones except for specific cases. I like using things such as humanizer and deep research, and it can be useful to take an existing skill and adjust it to your preferences. For example, my coding convention skill is edited from https://rworks.dev/posts/claude-skills-for-r-users/.

For tasks that are not obvious, I prefer to start in plan mode or ask the AI to restate the task before implementing. Sometimes I ask it to write the prompt back to me, then I edit the prompt and use it in a new chat. This can help avoid very long chats and makes unclear assumptions more visible. It is also useful to ask the AI to interview you when there are missing choices. For example:

```text
Bad: fix this merge

Better: use the project context and data-check instructions, fix the country-year merge, and report duplicate keys, unmatched rows, and any row-count changes
```

### AI and validation

AI is useful for writing code, explaining unfamiliar packages, and finding possible bugs, but plausible code can still do the wrong thing. I would ask it to add messages and checks, summarize what changed, inspect examples rather than only summary statistics, and compare important constructed variables to an external source when possible.


# Specific datasets

## WDPA (World Database on Protected Areas)

**Points** - The database includes protected areas (PAs) represented both as polygons and as points. For polygons, the boundaries of the PAs are available directly. Points are more ambiguous, as they can represent point-like features, or refer to larger protected areas for which the shape is unknown. The `REP_AREA` field can be used to approximate the size of these point records, but the point location is not necessarily the centroid of the area, and the shape of the area is unknown. Therefore, buffering points by their reported area may be useful as an approximation, but it is unlikely to recover the true shape of the PA.

**Overlaps** - When intersecting WDPA with other spatial data, for example to estimate the share of an area covered by PAs, need to account for overlapping of PAs. A given location can be covered by multiple PAs, so simply summing areas across records can overstate coverage. Thus, protected-area geometries may need to be unioned before calculating total covered area (perhaps by category, depending on the goal).

**Status year** - The time variable in WDPA is `STATUS_YR`, which records the last year of status change. This usually refers to the legal designation as a protected area. In checks against South African protected-area data from the late 1990s onward, most protected areas did not appear to change status after their initial designation during the observed period. This suggests that `STATUS_YR` can often be informative about designation timing, but it should still be used cautiously, especially for cases involving expansions, reclassification, proposed or established areas, and loss of status.

## IPUMS

# Suggestions for stuff to add / remove

- add a short opening paragraph about what kinds of RA work this is based on
- add one concrete example of a result validation workflow
- add a small checklist at the end
- maybe remove or shorten any examples that are too project-specific once the structure is clearer
